```{r data generation, echo = FALSE, results = "hide"}
XMean=5+sample(10,1)
YMean=-5+sample(10,1)
XVar=sample(5,1)
YVar=sample(5,1)
XSize=1+sample(3,1)
YSize=1+sample(3,1)
Const=sample(5,1)

## QUESTION/ANSWER GENERATION
Drop=1+sample(5,1)
questions <- character(5)
questions[1] <- paste(Const+XSize*XMean+YSize*YMean, "and", XSize^2*XVar+YSize^2*YVar, sep=" ")
questions[2] <- paste(Const+XSize*XMean+YSize*YMean, "and", XSize*XVar+YSize*YVar, sep=" ")
questions[3] <- paste(XSize^2*XMean+YSize^2*YMean, "and", XSize^2*XVar+YSize^2*YVar, sep=" ")
questions[4] <- paste(XMean+YMean, "and", XVar+YVar, sep=" ")
questions[5] <- paste(XSize*XMean+YSize*YMean, "and", XSize^2*XVar+YSize^2*YVar, sep=" ")
questions[6]=paste(Const+XSize*XMean+YSize*YMean, "and", XSize^2*XVar+YSize^2*YVar+Const, sep=" ")
questions=questions[-Drop]
solutions <- c(TRUE,FALSE,FALSE,FALSE,FALSE)
explanations <- character(5)
explanations[1] <- ""
explanations[2] <- ""
explanations[3] <- ""
explanations[4] <- ""
explanations[5] <- ""

## permute order of solutions/questions
Order <- sample(1:5)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

You are told that the random variable $X$ has a mean of $`r XMean`$ and a variance of $`r XVar`$, and is independent of the random variable $Y$ which has a mean of $`r YMean`$ and variance $`r YVar`$.

If $Z=`r XSize`X+`r YSize`Y + `r Const`$, what are the mean and variance of $Z$?

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

Recall that if $Z = aX + bY + c$ then $\mbox{E}(Z) = a \mbox{E}(X) + b \mbox{E}(Y) + c$ and, if $X$ and $Y$ are independent, $\mbox{var}(Z) = a^2 \mbox{var}(X) + b^2 \mbox{var}(Y)$. So the mean of $Z$ is <code>`r XSize` * (`r XMean`) + `r YSize` * (`r YMean`) + `r Const`</code>, and the variance is <code>`r XSize`^2 * (`r XVar`) + `r YSize`^2 * (`r YVar`)</code>.

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================
extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice
