```{r data generation, echo = FALSE, results = "hide"}
# created MBJ, confirmed JG
## DATA GENERATION
colors <- rmultinom(1, 100, c(0.2,0.3,0.2,0.2,0.1))/100
colors

## QUESTION/ANSWER GENERATION
questions <- numeric(5)
questions[1] <- round(colors[4]/(colors[4]+colors[2]),2)
questions[2] <- colors[4]
questions[3] <- colors[2]
questions[4] <- round(colors[4]/colors[2],2)
questions[5] <- round(colors[2]/colors[4],2)

# Check:
change <- which(abs(questions[2:5] - questions[1]) < 0.03) + 1
questions[change] <- questions + 0.05
questions <- as.character(questions)

solutions <- c(TRUE, FALSE, FALSE,FALSE,FALSE)
questions

explanations <- character(5)
explanations[1] <- ""
explanations[2] <- ""
explanations[3] <- ""
explanations[4] <- ""
explanations[5] <- ""

## permute order of solutions/questions
Order <- sample(1:5)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

The colors of m&m's chocolate candy follow the distribution given below.  Imagine that you are colour blind and can tell that a particular m&m is not yellow, orange, or brown, but cannot tell if it is red or green.  What is the probability that it is green?

```{r, echo=FALSE}
colors <- data.frame(matrix(colors, nrow=1))
names(colors) <- c("Yellow", "Red", "Orange", "Green", "Brown")
cbind(Colour = 'Probability', colors) |>
  knitr::kable(format='html', booktabs=TRUE, align='lrrrrr') |>
  kableExtra::kable_styling(bootstrap_options = "bordered", full_width = FALSE, position = "center")
```

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

We want $\Pr(green| \mbox{ either }red \mbox{ or } green)= \frac{\Pr(green)}{\Pr(green)+\Pr(red)}$. 

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================

extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice

