```{r data generation, echo = FALSE, results = "hide"}
## DATA GENERATION
RY = sample(3000,1)/10000
RG=sample(4700,1)/10000
AG = sample(1100,1)/10000
AY=1-RY-RG-AG
## QUESTION/ANSWER GENERATION
questions <- character(6)
    questions[1] <- AY
    questions[2] <- "1.0000"
    questions[3] <- 1+RY+RG+AG
    questions[4] <- RY+RG+AG
    questions[5] <- "0.0000"
    questions[6] <- RG+RY+AG-1

explanations <- character(6)
explanations[1] <- "Probabilities must sum to one."
explanations[2] <- "This would suggest that all peas were angular and yellow."
explanations[3] <- "Probabilities are less than or equal to one."
explanations[4] <- "This number should be subtracted from one."
explanations[5] <- "This would suggest that there are no angular and yellow peas."
explanations[6] <- "Probabilities cannot be negative!"

# drop one of  the wrong answers.
Drop=sample(5,1)+1
explanations <- explanations[-Drop]
questions <- questions[-Drop]
solutions <- c(TRUE, FALSE, FALSE, FALSE, FALSE)

## permute order of solutions/questions
Order <- sample(1:5)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

The seeds of a certain variety of peas can be classified into four distinct groups based on their appearance. These are round and yellow (RY), round and green (RG), angular and yellow (AY) and angular and green (AG). The table below shows the probability function for a given pea belonging to one of these four groups.

```{r, echo=FALSE}
peas <- data.frame(`Pea Type` = "P(Pea Type)", RY, RG, AY = '?', AG, check.names=FALSE)
peas |>
  knitr::kable(format='html', booktabs=TRUE, align='lrrrr') |>
  kableExtra::kable_styling(bootstrap_options = "bordered", full_width = FALSE, position = "center")
```

The probability of a pea being classified as angular and yellow is closest to:

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================

extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice

