```{r data generation, echo = FALSE, results = "hide"}
# created MBJ, confirmed JG
## DATA GENERATION
library(tidyverse)
colors <- rmultinom(1, 100, c(0.2,0.2,0.2,0.2,0.2))/100

## QUESTION/ANSWER GENERATION
questions <- character(5)
    questions[1] <- colors[5]
    questions[2] <- colors[5]+.1
    questions[3] <- colors[5]+.01
    questions[4] <- colors[5]-.1
    questions[5] <- colors[5]-.01
                 
    solutions <- c(TRUE, FALSE, FALSE,FALSE,FALSE)

explanations <- character(5)
explanations[1] <- ""
explanations[2] <- ""
explanations[3] <- ""
explanations[4] <- ""
explanations[5] <- ""

## permute order of solutions/questions
Order <- sample(1:5)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

The colors of m&m's chocolate candy follow the distribution given below.  What is the proportion of brown m&m's

```{r, echo=FALSE}
colors <- data.frame(matrix(colors, nrow=1))
names(colors) <- c("Yellow", "Red", "Orange", "Green", "Brown")
cbind(Colour = 'Probability', colors |> mutate(Brown = '?')) |>
  knitr::kable(format='html', booktabs=TRUE, align='lrrrrr') |>
  kableExtra::kable_styling(bootstrap_options = "bordered", full_width = FALSE, position = "center")
```

\begin{center}
\begin{tabular}{|c|c|c|c|c|c|} \hline
Color:& Yellow & Red & Orange & Green & Brown \\
\hline
Probability: & \Sexpr{colors[1]} & \Sexpr{colors[2]} &\Sexpr{colors[3]} & \Sexpr{colors[4]} & ?\\
\hline
\end{tabular}
\end{center}

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

Assuming we have all the colours covered, the probabilities must sum to 1 as the colours are mutually exclusive. So the probability of a Brown m\&m is one minus the sum of the other probabilities.

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================

extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice
