```{r data generation, echo = FALSE, results = "hide"}
## DATA GENERATION
AX= sample(5,1)+19
AY= sample(5,1)+18
AZ= sample(5,1)+27
AT=AX+AY+AZ
BX= sample(5,1)+14
BY= sample(5,1)+12
BZ= sample(5,1)+14
BT=BX+BY+BZ
CX= sample(5,1)+22
CY= sample(5,1)+19
CZ= sample(5,1)+33
CT=CX+CY+CZ
XT=AX+BX+CX
YT=AY+BY+CY
ZT=AZ+BZ+CZ
GT=AT+BT+CT

## QUESTION/ANSWER GENERATION
questions <- character(5)
    questions[1] <- round(XT/GT,4)
    questions[2] <- round(AT/XT,4)
    questions[3] <- round(BT/XT,4)
    questions[4] <- round(AT/GT,4)
    questions[5] <- round(GT/XT,4)
    solutions <- c(TRUE, FALSE, FALSE,FALSE,FALSE)

explanations <- character(5)
explanations[1] <- ""
explanations[2] <- ""
explanations[3] <- ""
explanations[4] <- ""
explanations[5] <- ""

## permute order of solutions/questions
Order <- sample(1:5)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

Joint problems are common in older dogs. To ease the problem, vets recommend a combination of certain types of food and supplements. In order to gauge the popularity of prescribed combination of food and supplement types, one study classified the options chosen by the owners of `r GT` randomly selected older dogs with joint problems, according to their food and supplement intake. The data are summarised in the following frequency table.

```{r, echo=FALSE}
foods <- data.frame(` ` = c(paste("Food", c("A", "B", "C")), "Total"), `Supplement X` = c(AX, BX, CX, XT), `Supplement Y` = c(AY, BY, CY, YT), `Supplement Z` = c(AZ, BZ, CZ, ZT), Total = c(AT, BT, CT, GT), check.names=FALSE)
foods |>
  knitr::kable(format='html', booktabs=TRUE, align='lrrrrr') |>
  kableExtra::kable_styling(bootstrap_options = "bordered", full_width = FALSE, position = "center") |>
  kableExtra::column_spec(column = 1, bold = TRUE)
```

If a dog is chosen at random, the probability that it will be on Supplement X is closest to:

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

Dogs are chosen at random, so the dogs to be considered are all dogs, or `r GT`. Of these, the number on Supplement X is `r XT`.

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================

extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice

