```{r data generation, echo = FALSE, results = "hide"}
## QUESTION/ANSWER GENERATION
questions <- character(5)
    questions[1] <- "0.004"
    questions[2] <- "0.001"
    questions[3] <- "0.002"
    questions[4] <- "0.003"
    questions[5] <- "0.005"

solutions <- c(TRUE,FALSE,FALSE,FALSE,FALSE)
explanations <- character(5)
explanations[1] <- ""
explanations[2] <- ""
explanations[3] <- ""
explanations[4] <- ""
explanations[5] <- ""

## permute order of solutions/questions
Order <- sample(1:5)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

It was amazing that Paul the octopus made eight correct predictions for the 2010 World Cup. Suppose that we were going to predict the 8 matches by tossing a balanced coin. Assuming that the tosses are independent, the probability of making eight correct predictions is closest to

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

The probability of getting one match correct by a balanced coin is 0.5. As the tosses are independent, the probability of making 8 correct predictions will be $0.5^8$. We could also use the binomial distribution for this, by using `dbinom(8,size=8,prob=0.5)`, or with the formula:

$$
P(X=8) = {8 \choose 8} 0.5^8 0.5^{0} = 0.5^8.
$$

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================

extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice
