```{r data generation, echo = FALSE, results = "hide"}
What=sample(c("car", "computer", "laptop", "television", "stereo"), 1)
HowMany= 5+sample(5,1)
Exposed=10+sample(20,1)
PEach = 1/Exposed

## QUESTION/ANSWER GENERATION
questions <- character(5)
    questions[1] <- round(dbinom(2, HowMany, PEach), 4)
    questions[2] <- round(pbinom(2, HowMany, PEach), 4)
    questions[3] <- 1-round(dbinom(2, HowMany, PEach), 4)
    questions[4] <- round(dbinom(2, Exposed, PEach), 4)
    questions[5] <- round(pbinom(2, Exposed, PEach), 4)

solutions <- c(TRUE,FALSE,FALSE,FALSE,FALSE)
explanations <- character(5)
explanations[1] <- ""
explanations[2] <- ""
explanations[3] <- ""
explanations[4] <- ""
explanations[5] <- ""

## permute order of solutions/questions
Order <- sample(1:5)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

A new `r What`  salesperson knows that they sell `r What`s to one out of every `r Exposed` customers. The probability that they will
sell a `r What` to exactly two of the next `r HowMany` customers is

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

The probability of two out of `r HowMany` is clearly a binomial probability question. We want exactly two, so we could use `dbinom` for this in RStudio. The probability of a person buying a `r What` is one out of `r Exposed`, so we can use <code>dbinom(2, size=`r HowMany`, prob=1/`r Exposed`)</code>.

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================

extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice
