```{r data generation, echo = FALSE, results = "hide"}
## QUESTION/ANSWER GENERATION
questions <- character(5)
    questions[1] <- "$0.8^{30}$."
    questions[2] <- "1."
    questions[3] <- "0.5."
    questions[4] <- "1/30."
    questions[5] <- "$0.2^{30}$."

solutions <- c(TRUE,FALSE,FALSE,FALSE,FALSE)
explanations <- character(5)
explanations[1] <- ""
explanations[2] <- ""
explanations[3] <- ""
explanations[4] <- ""
explanations[5] <- ""

## permute order of solutions/questions
Order <- sample(1:5)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

A student takes a statistics test with 30 multiple choice questions
where each  question  has five choices. Suppose that they just randomly
choose an answer for each question.  What is the probability that the
student gets zero answers correct?

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

The probability that they get zero answers correct is the same as the probability they get all 30 answers wrong. The probability they get an answer wrong by guessing is 4/5=0.8. Assuming independence (implied by guessing), the product of 0.8, 30 times is the answer. We could also use the binomial distribution for this:

$$
P(0) = {30 \choose 0} 0.2^0 0.8^{30} = 0.8^{30}
$$

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================

extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice
