```{r data generation, echo = FALSE, results = "hide"}
## QUESTION/ANSWER GENERATION
questions <- character(5)
    questions[1] <- "$\\bar{x}$ chart"
    questions[2] <- "$p$ chart"
    questions[3] <- "$c$ chart"
    questions[4] <- "$np$ chart"
    questions[5] <- "$u$ chart"

solutions <- c(TRUE,FALSE,FALSE,FALSE,FALSE)
explanations <- character(5)
explanations[1] <- "weights are continuous."
explanations[c(2,4)] <- "This is for binomially distributed counts."
explanations[c(3,5)] <- "This is for Poisson distributed counts."

## permute order of solutions/questions
Order <- sample(1:5)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

A new machine is placed into service at the New Zealand Mint to record 
the weights of new coins, which of the following control charts can be
used to directly monitor the weight of the coins?

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

Only the $\bar{x}$ chart is for monitoring continuous measures. The others ($p$, $np$, $c$, $u$) are for monitoring defects.

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================

extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice
