```{r data generation, echo = FALSE, results = "hide"}
library(exams)
## QUESTION/ANSWER GENERATION
questions <- character(5)
    questions[1] <- "$\\bar{x}$ chart."
    questions[2] <- "$p$ chart."
    questions[3] <- "$np$ chart."
    questions[4] <- "$c$ chart."
    questions[5] <- "$u$ chart."

solutions <- c(TRUE,FALSE,FALSE,FALSE,FALSE)
explanations <- character(5)

## permute order of solutions/questions
Order <- sample(1:5)
questions <- questions[Order]
solutions <- solutions[Order]
```

Question
========

Which of the following is not a control chart for an attribute?

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

Attributes control charts are for nonconformities or defects being counted. Only the distribution assumed changes among the four options ($p$, $np$, $c$, $u$) here. The $\bar{x}$ chart is a variables control chart which is used for continuous type measurements.

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================

extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice
