```{r data generation, echo = FALSE, results = "hide"}
library(exams)
What=sample(c("network cables", "memory cards", "USB cables", "LCD monitors"), 1)
file=paste0(unlist(strsplit(What, " "))[1], ".csv")
Rate=3.5+sample(10,1)/10
X=rpois(30,Rate)
Mean=mean(X)
Answer=Mean+3*sqrt(Mean)
dat = data.frame(Faults=X)
write.csv(dat, file, row.names = FALSE)

## QUESTION/ANSWER GENERATION
questions <- character(5)
questions[1] <- round(Answer,2)
questions[2] <- round(1.1*Answer,2)
questions[3] <- round(0.9*Answer,2)
questions[4] <- round(2*Answer,2)
questions[5] <- round(1/Answer,3)

solutions <- c(TRUE,FALSE,FALSE,FALSE,FALSE)
explanations <- character(5)
explanations[1] <- ""
explanations[2] <- ""
explanations[3] <- ""
explanations[4] <- ""
explanations[5] <- ""

## permute order of solutions/questions
Order <- sample(1:5)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

You are given the number of faulty `r What` in the last thirty shipments leaving the factory in the dataset `r file`.

Download the data file [`r file`](`r file`), and read it into RStudio.

By using the appropriate control chart for this kind of data, the upper control limit is closest to...

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

We could use RStudio for this with the `qcc` function with `type='c'`.

```{r gen1, echo=FALSE, message=FALSE, comment = "", results='asis', fig.show='hide'}
src <- knitr::knit_expand(text=c('```{r gend, message=FALSE, fig.dim=c(7,5), fig.cap="", results="hide"}',
                   'library(tidyverse)',
                   'library(qcc)',
                   'dat <- read_csv("{{file}}")',
                   'qcc(dat$Faults, type="c")',
                   '```'))
res = knitr::knit_child(text = src, options = list(comment = ""), quiet=TRUE)
cat(res)
```

Alternatively, we could use the formula: $UCL = \bar{c} + 3 \sqrt{\bar{c}}$.

```{r ans2, message=FALSE}
library(tidyverse)
dat |> summarise(c = mean(Faults),
                  ucl = c + 3 * sqrt(c))
```

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================

extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice
