```{r data generation, echo = FALSE, results = "hide"}
## QUESTION/ANSWER GENERATION
questions <- character(5)
questions[1] <- "Normal distribution"
questions[2] <- "Binomial distribution"
questions[3] <- "Gamma distribution"
questions[4] <- "Exponential distribution"
questions[5] <- "Poisson distribution"

solutions <- c(TRUE,FALSE,FALSE,FALSE,FALSE)
explanations <- character(5)
explanations[1] <- ""
explanations[2] <- ""
explanations[3] <- ""
explanations[4] <- ""
explanations[5] <- ""

## permute order of solutions/questions
Order <- sample(1:5)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

Unquestionably, the most important and useful
distribution in probability and statistics is the

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

The Central Limit Theorem shows that regardless of population distribution, the distribution
of the sample mean follows a normal distribution. This is what makes the Normal so useful!

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================

extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice
