```{r data generation, echo = FALSE, results = "hide"}
library(exams)
## DATA GENERATION
## QUESTION/ANSWER GENERATION
questions <- character(5)
questions[1] <- "Events $A$ and $B$ are not independent."
questions[2] <- "There is not enough information to judge the independence/dependence between events $A$ and $B$."
questions[3] <- "Events $A$ and $B$ are independent."
questions[4] <- "The conditional probability $P(B|A)>0$."
questions[5] <- "The conditional probability $P(A|B)>0$."

solutions <- c(TRUE,FALSE,FALSE,FALSE,FALSE)
explanations <- character(5)
explanations[1] <- "Yes, we know they cannot be independent."
explanations[2] <- "We know $P(A \\mbox{ and } B) \\neq P(A) \\times P(B)$."
explanations[3] <- "We know $P(A \\mbox{ and } B) \\neq P(A) \\times P(B)$."
explanations[4] <- "$P(B|A) = \\frac{P(B \\mbox{ and } A)}{P(A)} = 0$."
explanations[5] <- "$P(A|B) = \\frac{P(A \\mbox{ and } B)}{P(B)} = 0$."

## permute order of solutions/questions
Order <- sample(1:5)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]
```

Question
========

Two events, say event $A$ and event $B$, are disjoint (i.e. $A$ and
$B$ do not have common outcomes), and neither of
them is an impossible event (i.e.
$P(A) \neq 0$ and $P(B) \neq 0$. Which following statement is correct?

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

The two events are disjoint, so they cannot both occur so $P(A \mbox{ and } B) = 0$. If two events are independent then $P(A \mbox{ and } B) = P(A)P(B) = 0$, but in this case we know  $P(A) \neq 0$ and $P(B) \neq 0$ so that $P(A)P(B) \neq 0$.

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================

extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice
