```{r data generation, echo = FALSE, results = "hide"}
library(exams)
## DATA GENERATION
## QUESTION/ANSWER GENERATION
questions <- character(5)
questions[1] <- "The two events must not be independent."
questions[2] <- "It is possible that $P(A) + P(B) > 1$."
questions[3] <- "It is possible that $P(A) \\times P(B) < 0$."
questions[4] <- "The two events must be independent."
questions[5] <- "The two events can be independent or dependent, depending on the probabilities of the two events."

solutions <- c(TRUE,FALSE,FALSE,FALSE,FALSE)
explanations <- character(5)
explanations[1] <- "Yes, the events must be dependent."
explanations[2] <- "If $P(A) + P(B) > 1$, then we must have $P(A \\mbox{ and } B) > 0$, as $P(A \\mbox{ or } B) = P(A) + P(B) - P(A \\mbox{ and } B) \\le 1$ as it is a probability."
explanations[3] <- "$P(A) \\ge 0$ and $P(B) \\ge 0$ so that $P(A) \\times P(B) \\ge 0$."
explanations[4] <- "No, the events must be dependent"
explanations[5] <- "No, the events must be dependent"

## permute order of solutions/questions
Order <- sample(1:5)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```


Question
========

Suppose that events $A$ and $B$ are disjoint with $P(A) \ne 0$,and $P(B) \ne 0$. Which statement is correct?

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

As $A$ and $B$ are disjoint, we know that $P(A \mbox{ and } B) = 0$, as they can't both occur. To be
independent, we have $P(A \mbox{ and } B) = P(A) \times P(B) = 0$, so this means that one of $P(A)$ or $P(B)$ must
be zero, but we know they're not. Thus, they must be dependent.

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================

extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice
