```{r data generation, echo = FALSE, results = "hide"}
library(exams)

Mean=1900 + sample(10,1)*20
HisCount=2900 + sample(10,1)*20
## QUESTION/ANSWER GENERATION
questions <- character(5)
questions[1] <- (HisCount-Mean)/2
questions[2] <- HisCount-Mean
questions[3] <- round((HisCount-Mean)/6,0)
questions[4] <- round((HisCount-Mean)/4,0)
questions[5] <- round((HisCount-Mean)/3,0)

solutions <- c(TRUE,FALSE,FALSE,FALSE,FALSE)
explanations <- character(5)
explanations[1] <- ""
explanations[2] <- ""
explanations[3] <- ""
explanations[4] <- ""
explanations[5] <- ""

## permute order of solutions/questions
Order <- sample(1:5)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

A man's blood test shows a cholesterol count of `r HisCount`mg/l. If the doctor says that the average cholesterol count for men of
a similar age is `r Mean`mg/l and that only about 2.5% of the male
population will have a value higher than `r HisCount`mg/l, the standard
deviation assuming the cholesterol counts are normally distributed is
closest to...

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

We can use the rule of thumb of the normal distribution, noting that 95% of observations are within 2 standard deviations of the mean, which leaves 2.5% above this level. Thus we must have that $`r HisCount` = `r Mean` + 2 \times \sigma$, which we can solve for $\sigma$.

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================
extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice
