```{r data generation, echo = FALSE, results = "hide"}
HowMany=100+10*sample(10,1)
Mean=22+sample(6,1)
SD=3+sample(10,1)/2

## QUESTION/ANSWER GENERATION
questions <- character(5)
    questions[1] <- paste("About ", round(0.68*HowMany,0), " of the students would have scores between ", Mean-SD, " and ", Mean+SD, ".", sep="")
    questions[2] <- paste("About ", round(2*HowMany/3,0), " of the students would have scores between ", Mean-2*SD, " and ", Mean+2*SD, ".", sep="")
    questions[3] <- paste("About ", round(0.95*HowMany,0), " of the students would have scores between ", Mean-SD, " and ", Mean+SD, ".", sep="")
    questions[4] <- paste("About ", round(0.8*HowMany,0), " of the students would have scores between ", Mean-SD, " and ", Mean+SD, ".", sep="")
    questions[5] <- paste("About ", round(0.99*HowMany,0), " of the students would have scores between ", Mean-2*SD, " and ", Mean+2*SD, ".", sep="")
    solutions <- c(TRUE,FALSE, FALSE, FALSE,FALSE)

explanations <- character(5)
explanations[1] <- ""
explanations[2] <- ""
explanations[3] <- ""
explanations[4] <- ""
explanations[5] <- ""

## permute order of solutions/questions
Order <- sample(1:5)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

The Chapman/Brown test is a psychological test designed to measure how people react to stressful situations, like mastery tests for example. The Chapman/Brown test was given to a group of `r HowMany` university students as a trial. It was reported that the mean score in the test was `r Mean` and the standard deviation was `r SD`. Assuming the individual test scores were normally distributed, which of the following statements is most correct?

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

Recall the rules of thumb for the normal distribution: 68% of observations are within one, 95% within two, and 99.5% within three standard deviations of the mean. Given we have `r HowMany` students, this means we'd expect `r round(0.68*HowMany,0)` students within one, `r round(0.95*HowMany,0)` within two, and `r round(0.995*HowMany,0)` within three standard deviations of the mean.

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================
extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice
