```{r data generation, echo = FALSE, results = "hide"}

questions <- character(5)
questions[1] <- "Less than 1%."
questions[2] <- "Less than 2.5%."
questions[3] <- "More than 5%, but less than 10%."
questions[4] <- "Approximately half of them."
questions[5] <- "0% (None), as the only men who get a score over 2500
will be unhealthy and should be put on medication as soon as possible."

solutions <- c(TRUE,FALSE,FALSE,FALSE,FALSE)
explanations <- character(5)
explanations[1] <- ""
explanations[2] <- ""
explanations[3] <- ""
explanations[4] <- ""
explanations[5] <- ""

## permute order of solutions/questions
Order <- sample(1:5)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

The average cholesterol for healthy men is 1900, with a standard deviation of 200. What proportion of males
will have a cholesterol count more than 2500?

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

Noting that 2500 is three standard deviations from 1900, we can use the rule of thumb from the normal distribution which tells us that 99.5% of observations are within three standard deviations of the mean, so around 0.5% are outside of this. Thus around 0.25% are at each end.

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================
extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice
