```{r data generation, echo = FALSE, results = "hide"}
Mean=8+sample(5,1)
SD=2+sample(5,1)
## QUESTION/ANSWER GENERATION
questions <- character(5)
    questions[1] <- round( pnorm(16, Mean, SD) -pnorm(6, Mean, SD), 4)
    questions[2] <- round( pnorm(16, Mean, SD) + pnorm(6, Mean, SD), 4)
    questions[3] <- round( pnorm(16, Mean, SD), 4)
    questions[4] <- 1 - round( pnorm(16, Mean, SD) -pnorm(6, Mean, SD), 4)
    questions[5] <- 1- round( pnorm(6, Mean, SD), 4)

solutions <- c(TRUE,FALSE,FALSE,FALSE,FALSE)
explanations <- character(5)
explanations[1] <- ""
explanations[2] <- ""
explanations[3] <- ""
explanations[4] <- ""
explanations[5] <- ""

## permute order of solutions/questions
Order <- sample(1:5)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```


Question
========

If $X$ is a normally distributed random variable with a mean of `r Mean` and standard deviation of `r SD`, then the
probability that $X$ is between 6 and 16 is 

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

We can use two applications of `pnorm()` for this, recalling that <code>pnorm(x, mean=`r Mean`, sd=`r SD`)</code> returns $P(X < x)$,
so that $P(6 < X < 16)$ can be computed using <code>pnorm(16, mean=`r Mean`, sd=`r SD`) - pnorm(6, mean=`r Mean`, sd=`r SD`)</code>.

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================
extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice
