```{r data generation, echo = FALSE, results = "hide"}
Gender=sample(c("", "s"), 1)
Mean=40+sample(10,1)
SD=3+sample(5,1)
Product=sample(c("book", "DVD", "perfume"), 1)

## QUESTION/ANSWER GENERATION
questions <- character(5)
    questions[1] <- ceiling(Mean+1.645*SD)
    questions[2] <- ceiling(Mean+1.96*SD)
    questions[3] <- floor(Mean+3*SD)
    questions[4] <- ceiling(Mean+SD)
    questions[5] <- ceiling(Mean+2.5*SD)

solutions <- c(TRUE,FALSE,FALSE,FALSE,FALSE)
explanations <- character(5)
explanations[1] <- ""
explanations[2] <- ""
explanations[3] <- ""
explanations[4] <- ""
explanations[5] <- ""

## permute order of solutions/questions
Order <- sample(1:5)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

A retailer finds that the demand for a very popular
`r Product` averages `r Mean` per week with a standard deviation of `r SD`. If new stock arrives exactly one week after an order is placed, and the seller wishes to have adequate stock 95% of the time, the number
of `r Product`s `r Gender`he must keep on hand is approximately...

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

The seller wishes to have adequate stock on hand 95% of the time. So they need to know the value $X$ such that $P(sales < X) = 0.95$. As sales follows a normal distribution, we can use `qnorm` to find this 95th quantile via <code>qnorm(0.95, mean=`r Mean`, sd=`r SD`)</code>.

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================
extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice


\begin{question}

%%%make no adjustments past this point!!!
\begin{answerlist}
  \item \Sexpr{questions[1]}
  \item \Sexpr{questions[2]}
  \item \Sexpr{questions[3]}
  \item \Sexpr{questions[4]}
  \item \Sexpr{questions[5]}
\end{answerlist}
\end{question}

%% SOLUTIONS
\begin{solution}
\begin{answerlist}
  \item \Sexpr{mchoice2text(solutions[1])}: \Sexpr{explanations[1]}
  \item \Sexpr{mchoice2text(solutions[2])}: \Sexpr{explanations[2]}
  \item \Sexpr{mchoice2text(solutions[3])}: \Sexpr{explanations[3]}
  \item \Sexpr{mchoice2text(solutions[4])}: \Sexpr{explanations[4]}
  \item \Sexpr{mchoice2text(solutions[5])}: \Sexpr{explanations[5]}
\end{answerlist}
\end{solution}

%% META-INFORMATION
%% \extype{schoice}
%% \exsolution{\Sexpr{mchoice2string(solutions)}}
%% \exname{Multiple choice}

