```{r data generation, echo = FALSE, results = "hide"}
library(masseystat)
# created MBJ, confirmed JG

lambda<-round(abs(rnorm(1,3,2)),2)
x<-rpois(1, lambda)+c(1,3)

## QUESTION/ANSWER GENERATION
questions <- numeric(5)
questions[1] <-  round(sum(dpois(seq(x[1],x[2],1),lambda)),4)
questions[2] <-  round(dpois(x[1], lambda), 4)
questions[3] <-  round(ppois(x[1], lambda), 4)
questions[4] <-  round(dpois(x[2], lambda)+0.05,4)
questions[5] <- round(ppois(x[2],lambda)-ppois(x[1],lambda), 4) 

solutions <- c(TRUE,FALSE,FALSE,FALSE,FALSE)
questions <- make_unique(questions, solutions, tolerance=0.01)
questions <- as.character(questions)

explanations <- character(5)
explanations[1] <- ""
explanations[2] <- ""
explanations[3] <- ""
explanations[4] <- ""
explanations[5] <- ""

## permute order of solutions/questions
Order <- sample(1:5)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

$X$ has a Poisson distribution with rate `r lambda`.  Find the probability that $`r x[1]` \le X \le `r x[2]`$.

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

$P(`r x[1]` \le X \le `r x[2]`) = P(X \le `r x[2]`) - P(X < `r x[1]`)$. We can use <code>ppois(`r x[2]`, lambda=`r lambda`)</code> for the first part, and <code>ppois(`r x[1]-1`, lambda=`r lambda`)</code> for the second (remembering that `ppois` includes the equal to case).

Alternatively, we could use <code>sum(dpois(`r x[1]`:`r x[2]`, lambda=`r lambda`))</code>.

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================

extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice
