```{r data generation, echo = FALSE, results = "hide"}
library(exams)
# created MBJ, JG says too hard for MT1, used in MT3

## DATA GENERATION
Which=sample(3,1)
What=c("wine", "beer", "whisky")[Which]

## QUESTION/ANSWER GENERATION
questions <- character(5)
    questions[1] <- paste("No glasses of", What, "are correctly assigned")
    questions[2] <- "All glasses are correctly assigned"
    questions[3] <- "Exactly two glasses are correctly assigned"
    questions[4] <- paste("The third and fourth glasses of", What, "tasted are correctly assigned")
    questions[5] <-  paste("The third and fourth glasses of", What, "tasted are incorrectly assigned")
    solutions <- c(TRUE, FALSE, FALSE,FALSE,FALSE)

explanations <- character(5)
explanations[1] <- "Yes, we can't both correctly assign two glasses while incorrectly assigning all glasses."
explanations[2] <- "No, the events 'two correct' and 'all correct' can both occur."
explanations[3] <- "No, the events 'two correct' and 'exactly two correct' can both occur."
explanations[4] <- "No, the event describing the assignment of the third and fourth glasses can occur alongside the event describing the assignment of the first two glasses."
explanations[5] <- "No, the event describing the assignment of the third and fourth glasses can occur alongside the event describing the assignment of the first two glasses."

## permute order of solutions/questions
Order <- sample(1:5)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

A panel of `r What` enthusiasts are given 4 labelled bottles, and 4
unlabelled glasses of `r What`. They are asked to match each `r What` with the
correct bottle. Consider the event that the first two glasses tasted are
correctly assigned to their bottles. Which of the following events are
mutually exclusive with (disjoint from) this event?


```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

Mutually exclusive means two events cannot happen at the same time.

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================

extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice