```{r data generation, echo = FALSE, results = "hide"}
# created MBJ, JG likes but not a basic question for MT1

library(tidyverse)
library(exams)
## DATA GENERATION
probs<-rbeta(5,0.5, 0.5)
probs[4]<- 0
probs[5]<-0.1
probs[1:3]<- 0.9*probs[1:3]/sum(probs[1:3])
probs<- round(probs,2)
probs[5]<- 1-sum(probs[1:3])
answer<-round(probs[5]/(sum(probs[2:5])),4)
## QUESTION/ANSWER GENERATION
questions <- character(5)
    questions[1] <- answer
    questions[2] <-round( answer*0.5,4)
    questions[3] <- answer-0.02
    questions[4] <- answer+0.02
    questions[5] <- answer*2
    solutions <- c(TRUE, FALSE, FALSE,FALSE,FALSE)

explanations <- character(5)
explanations[1] <- ""
explanations[2] <- ""
explanations[3] <- ""
explanations[4] <- ""
explanations[5] <- ""

## permute order of solutions/questions
Order <- sample(1:5)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

A panel of wine enthusiasts are given 4 labeled bottles, and 4 unlabeled glasses of wine. The are asked to match each wine with the correct bottle. The following table gives the probability of each number of correct assignments for this population of experts.

```{r, echo=FALSE}
probs <- data.frame(matrix(probs, nrow=1))
names(probs) <- 0:4
cbind(`Correct matches` = "Probability", probs) |>
  knitr::kable(format='html', booktabs=TRUE) |>
  kableExtra::kable_styling(bootstrap_options = "bordered", full_width = FALSE, position = "center")
```

If we restrict ourselves to experts that get at least one wine correct, what is the probability that they get all wines correct?

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

We want $\Pr(matches=4| matches \ge 1)= \frac{\Pr(matches=4)}{\Pr(matches=1)+\Pr(matches=2)+\Pr(matches=3)+\Pr(matches=4)}$. 

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================

extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice
