```{r data generation, echo = FALSE, results = "hide"}
library(exams)
PA=0.85+sample(10,1)/100
PB=0.92+sample(7,1)/100

## QUESTION/ANSWER GENERATION
questions <- character(5)
questions[1] <- round(PA*PB,4)
questions[2] <- round((PA+PB)/2, 4)
questions[3] <- round((1-PA)*(1-PB), 4)
questions[4] <- round(1-(1-PA)*(1-PB), 4)
questions[5] <- max(PA,PB)

solutions <- c(TRUE,FALSE,FALSE,FALSE,FALSE)
explanations <- character(5)
explanations[1] <- ""
explanations[2] <- ""
explanations[3] <- ""
explanations[4] <- ""
explanations[5] <- ""

## permute order of solutions/questions
Order <- sample(1:5)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

A system has two components, $A$ and $B$, which are connected in
series. The probability that component $A$ works is `r PA`, and the
probability that component $B$ works is `r PB`. Suppose that the
components work independently. What is the probability that the system
works?

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

As the components are in series, the system will only work if both components are working, so that $P(\mbox{System}) = P(A\mbox{ and }B)$. As $A$ and $B$ are independent,

$$
\begin{aligned}
P(\mbox{System}) = P(A\mbox{ and }B) &= P(A) \times P(B)\\
&= `r PA` \times `r PB`.
\end{aligned}
$$

In RStudio:

```
PA = `r PA`
PB = `r PB`
PA * PB
```

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================

extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice
