```{r data generation, echo = FALSE, results = "hide"}
library(exams)
PA=0.85+sample(10,1)/100
PB=0.92+sample(7,1)/100

## QUESTION/ANSWER GENERATION
questions <- character(5)
questions[1] <- round(1-(1-PA)*(1-PB), 4)
questions[2] <- round((PA+PB)/2, 4)
questions[3] <- round((1-PA)*(1-PB), 4)
questions[4] <- round(PA*PB,4)
questions[5] <- max(PA,PB)

solutions <- c(TRUE,FALSE,FALSE,FALSE,FALSE)
explanations <- character(5)
explanations[1] <- ""
explanations[2] <- ""
explanations[3] <- ""
explanations[4] <- ""
explanations[5] <- ""

## permute order of solutions/questions
Order <- sample(1:5)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

A system has two components, $A$ and $B$, which are connected in parallel. The probability that component $A$ works is `r PA`, and the probability that component $B$ works is `r PB`. If the components work independently of one another, what is the probability that the system works?

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

As the components are in parallel, the system will work if either $A$ or $B$ are working, so that $P(\mbox{System}) = P(A\mbox{ or }B)$. As the components are independent, we know that $P(A\mbox{ and } B) = P(A) \times P(B)$, so that

$$
\begin{aligned}
P(\mbox{System}) = P(A\mbox{ or }B) &= P(A) + P(B) - P(A) \times P(B)\\
& = `r PA` + `r PB` - `r PA` \times `r PB`.
\end{aligned}
$$

The other way to work this out is to recognise that the system will **not** work only if both $A$ and $B$ are not working. So $P(\mbox{System broken}) = P(A^c\mbox{ and }B^c)$, where $A^c$ is the complement of $A$ (i.e. $A^c$ means $A$ not working). Again, as $A$ and $B$ are independent, $P(A^c\mbox{ and } B^c) = P(A^c) \times P(B^c)$ so that

$$
\begin{aligned}
P(\mbox{System broken}) = P(A^c\mbox{ and }B^c) &= P(A^c) \times P(B^c)\\
&= (1 - P(A)) \times (1 - P(B))\\
&= (1 - `r PA`) \times (1 - `r PB`)
\end{aligned}
$$

Note that a little algebra will show that these are the same. In Rstudio:

```
Afails = 1 - `r PA`
Bfails = 1 - `r PB`
SystemFails = Afails * Bfails
1 - SystemFails
```

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================

extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice
