```{r data generation, echo = FALSE, results = "hide"}
HowMany=30+sample(20,1)
Mean= 20+sample(10,1)
SD=3+sample(5,1)
XPoint= as.integer(round(Mean+1.5*SD/sqrt(HowMany), 0))
## QUESTION/ANSWER GENERATION
questions <- character(5)
questions[1] <- round(1-pnorm(XPoint, Mean, SD/sqrt(HowMany)), 4)
questions[2] <- "Unknown because the distribution of the population is unknown."
questions[3] <- round(1-pnorm(XPoint, Mean, SD), 4)
questions[4] <- round(1-pnorm(XPoint, Mean, SD/HowMany), 4)
questions[5] <- "0.0500"

solutions <- c(TRUE,FALSE,FALSE,FALSE,FALSE)
explanations <- character(5)
explanations[1] <- ""
explanations[2] <- ""
explanations[3] <- ""
explanations[4] <- ""
explanations[5] <- ""

## permute order of solutions/questions
Order <- sample(1:5)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

A random sample of $n=`r HowMany`$
measurements is selected from a population with mean equal to `r Mean` and
standard deviation `r SD`. What is the probability that the sample mean will
exceed `r XPoint`?

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

Recall that the sample mean is distributed as $\mbox{Normal}(\mu, \frac{\sigma}{\sqrt{n}})$. For the probability greater than a value, we can use `pnorm` with `lower.tail=FALSE`. So in RStudio, we can use <code>pnorm(`r XPoint`, mean=`r Mean`, sd=`r SD`/sqrt(`r HowMany`), lower.tail=FALSE)</code>.

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================
extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice
