```{r data generation, echo = FALSE, results = "hide"}
SMean=4+sample(20,1)/10
Sigma=2.5+sample(10,1)/10
N=30+sample(10,1)
What=SMean+sample(10,1)/10

## QUESTION/ANSWER GENERATION
questions <- character(5)
questions[1] <- round(pnorm(What, SMean, Sigma/sqrt(N)), 4)
questions[2] <- round(1-pnorm(What, SMean, Sigma/sqrt(N)), 4)
questions[3] <- round(pnorm(What, SMean, Sigma), 4)
questions[4] <- round(pnorm(What, SMean, Sigma/N), 4)
questions[5] <- round(0.5*pnorm(What, SMean, Sigma/sqrt(N)), 4)

solutions <- c(TRUE,FALSE,FALSE,FALSE,FALSE)
explanations <- character(5)
explanations[1] <- ""
explanations[2] <- ""
explanations[3] <- ""
explanations[4] <- ""
explanations[5] <- ""

## permute order of solutions/questions
Order <- sample(1:5)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

Suppose we do know that the  population mean $\mu=`r SMean`$, the population standard deviation $\sigma=`r Sigma`$, and the sample is of size $n = `r N`$. The
probability that the sample mean $\bar X$ is less 
than `r What`  is closest to...

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

The sample mean is distributed as $\mbox{Normal}(\mu, \frac{\sigma}{\sqrt{n}})$. For a normal probability less than a value, we can use `pnorm`. So in RStudio, we can use <code>pnorm(`r What`, mean=`r SMean`, sd=`r Sigma`/sqrt(`r N`))</code>.

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================
extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice
