```{r data generation, echo = FALSE, results = "hide"}
library(exams)
SampleSize=sample(100,1)+50
SampleMean=sample(100,1)+60
SampleSD=sample(10,1)+20
## QUESTION/ANSWER GENERATION
questions <- character(5)
questions[1] <- round(SampleSD/sqrt(SampleSize), 2)
questions[2] <- round(SampleSD, 2)
questions[3] <- round(sqrt(SampleSD/SampleSize), 2)
questions[4] <- round(SampleSD^2/(SampleSize), 2)
questions[5] <- round(SampleSD/SampleSize, 2)
solutions <- c(TRUE, FALSE, FALSE, FALSE, FALSE)
explanations = character(5)

## permute order of solutions/questions
Order <- sample(1:5)
questions <- questions[Order]
solutions <- solutions[Order]
```

Question
========

Let $x_1,x_2, \ldots ,x_{`r SampleSize`}$ be a random sample from a population with mean of `r SampleMean` and a
standard deviation of `r SampleSD`. Then the standard deviation of $\bar{x}$ is closest to...

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

The standard error is the standard deviation `r SampleSD` divided by the square root of the sample size, so we could use <code>`r SampleSD`/sqrt(`r SampleSize`)</code>.

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================
extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice
