---
title: "Lecture C07"
date: "Time series analysis"
output:
  xaringan::moon_reader:
    css: [default, default-fonts, "part_c.css"]
    # mathjax: "https://cdn.bootcss.com/mathjax/2.7.3/MathJax.js?config=TeX-MML-AM_CHTML"
    nature:
      highlightStyle: tomorrow
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
editor_options:
  chunk_output_type: console
---


```{r setup, echo=FALSE, message=FALSE}
library(janitor)
library(knitr)
library(ggplot2); theme_set(theme_bw(base_size=15))
library(patchwork)
library(fontawesome)
library(visreg)
opts_chunk$set(#dev.args=list(bg='transparent'), 
               comment="", warning=FALSE, echo=FALSE)
knitr::opts_knit$set(root.dir = "../")
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.dim=c(8,5), out.width="70%", fig.retina=2)
```

## Learning Outcomes

- Time Series 

- Run chart

- Time Series Components

  - Trend 
  
  - Seasonality
  
  - Cycle
  
  - Irregularity

---

## Recap

- Simple linear model

- Inference and diagnostics

- Multivariate linear model

- Factors and interactions


---

class: middle, inverse

# Time Series

---


## What do we see over time?

--

Data obtained from observations collected sequentially over time are extremely common.

--

In business, we observe weekly interest rates, daily closing stock prices, monthly price indices, yearly sales figures, and so forth. 

--

In meteorology, we observe daily high and low temperatures, annual precipitation and drought indices, and hourly wind speeds. 

--

In agriculture, we record annual figures for crop and livestock production, soil erosion, and export sales. 

--

In the biological sciences, we observe the electrical activity of the heart at millisecond intervals. 

--

In ecology, we record the abundance of an animal species.

---


## Why do we study this? 

--

The list of areas in which time series are studied is virtually endless. 

--

The purpose of time series analysis is generally twofold: 

--

- to understand or model the stochastic mechanism that gives rise to an observed series and
    
- to predict or forecast the future values of a series based on the history of that series and, possibly, other related series or factors.

--
    
These are generally the same goals as for linear models - inference and prediction.

---


## Fundamental concepts

--

A **time series** is a series of data points indexed (or listed or graphed) in time order observed at successive equally spaced points in time. 

--

A **time series**  can be written as $Y_1,Y_2,\ldots,Y_T$ which is an ordered sequence of random variables.

--

The observations are made at times $t = 1,2,\ldots,T$.

--

A somewhat unique feature of time series and their models is that we usually cannot assume that the observations arise independently from a common population. 

--

Time series models incorporate various types of dependence among samples.

---

## Run Chart - Total value of NZ housing stock 

A **run chart** is a simple plot of time series observations over time.

```{r,echo=FALSE, out.width = "65%", message = FALSE}
# data downloaded from https://www.rbnz.govt.nz/statistics/series/economic-indicators/housing

read_rnzb <- function(file, skippy = 0) {
  require(readxl)
  require(janitor)
  read_xlsx(file, skip=5, 
              col_names = c("Date", read_xlsx(file, n_max=0, skip=skippy) %>% names ) ) %>%
  clean_names
}

housing <- read_rnzb("data/hm10.xlsx")

housing %>% 
  ggplot() + aes(x=date,y=total_value_of_housing_stock) + 
  geom_point() + geom_line() +
  xlab('Time') + ylab('Total value of NZ housing stock') + 
  ggtitle('Is the housing market always prospering?') +
  scale_y_continuous(labels = scales::dollar_format(prefix="$", scale=1/1e6, suffix = "T", accuracy=.5))
```

.small[
https://www.rbnz.govt.nz/statistics/series/economic-indicators/housing
]

---


## Run Chart - NZD/USD exchange rate

```{r,echo=FALSE}

exchangerate <- read_rnzb("Data/hb1-daily.xlsx", skippy=1) %>%
  bind_rows(read_rnzb("Data/hb1-daily-1999-2017.xlsx", skippy=1)) %>%
  bind_rows(read_rnzb("Data/hb1-daily-1973-1998.xlsx", skippy=1))

exchangerate %>% select(date, united_states_dollar, chinese_renminbi) %>% drop_na() %>%
  ggplot(aes(x=date, y=united_states_dollar)) +  geom_line() + 
  xlab('Time') + ylab('NZD to USD Exchange Rate') + 
  ggtitle('Is the value of the NZ dollar increasing over time?')
```

.small[
https://www.rbnz.govt.nz/statistics/series/exchange-and-interest-rates/exchange-rates-and-the-trade-weighted-index
]
---

## Run Chart - NZD/USD exchange rate

```{r,echo=FALSE}

exchangerate %>% select(date, united_states_dollar) %>% drop_na() %>%
  ggplot(aes(x=date, y=united_states_dollar)) +  geom_line() + 
  xlab('Time') + ylab('NZD to USD Exchange Rate') + 
  ggtitle('What if we look further back?')
```

.small[
https://www.rbnz.govt.nz/statistics/series/exchange-and-interest-rates/exchange-rates-and-the-trade-weighted-index
]
---

## Run Chart - NZ population

```{r,echo=FALSE, message=FALSE}
population <- read_rnzb("Data/hm12.xlsx") %>%
  rename(
    "resident_population" = "estimated_resident_population",
    "passenger_arrivals" = "passenger_travel_by_direction_3",
    "passenger_departures" = "passenger_travel_by_direction_4",
    "migration_arrivals" = "estimated_migration_by_direction_5",
    "migration_departures" = "estimated_migration_by_direction_6",
    "migration_net" = "estimated_migration_by_direction_7",
    "short_term_arrivals" = "estimated_short_term_travel_by_direction_8",
    "short_term_departures" = "estimated_short_term_travel_by_direction_9"
  )


gp1 <- population %>% 
  select(date, resident_population  ) %>% 
  drop_na() %>%
  ggplot(aes(x=date, y=resident_population)) + 
    geom_point() + geom_line() +
    xlab('Time') + ylab('NZ resident population') +
    scale_y_continuous(labels = scales::unit_format(unit = "M", 
                                                    scale = 1e-6),
                       limits = c(3400000,5500000))

gp1

```

.small[
https://www.rbnz.govt.nz/statistics/series/economic-indicators/population-and-migration
]
---

## Run Chart - NZ exports

```{r,echo=FALSE, message=FALSE}
overseastrade <- read_rnzb("Data/hm8.xlsx") %>%
  transmute(date = date,
            export_values = export_values_9 )

overseastrade %>% 
  ggplot(aes(x=date, y=export_values)) + 
  geom_point() + geom_line() + 
  scale_y_continuous(labels = scales::dollar_format(prefix="$", 
                                                    scale=1/1000, 
                                                    suffix = "B", 
                                                    accuracy=1)) + 
  xlab('Time') + ylab('Export value per quarter') + ggtitle('Goods sold by kiwis to the world') 

```

.small[
https://www.rbnz.govt.nz/statistics/series/economic-indicators/overseas-trade
]

---

class: middle, inverse

# Components of time series

---

## Components of time series

--

A time series may contain one or more of the following four components: 

  - trend, 
  - seasonal, 
  - cyclical, and 
  - irregular (unobserved or remainder)

--

These components, explained over the next few slides, describe how a time series behaves over time $t$.

--

The text book "Statistics for Business and Economics" by Prem S. Mann (1995), published by John Wiley,  gives a gentle introduction to the various components of a time series. 


---

class: middle, inverse

# Trend component

---

## Trend component

--

'Trend' is defined as the long-term movement (increase or decrease) in a time series. 
 
--

In other words, a trend is an increase or decrease in the values of a variable occurring over a
period of several years gives a trend.

--

If the values of a time series do not change on average over time, then that time series has no trend.
 
--

A trend shows the long-run pattern in the growth or decline in a time series that may represent a variable such as sales, inventories, revenues, profits, employment, population, housing starts, gross national product, consumer expenditure, investment, interest rates, etc.

---

## Some trends over time

```{r,echo=FALSE, fig.dim=c(12, 6), out.width="90%"}
set.seed(2020)
T<-24
freq <- 4
seasonalindex <- rep(scale(rdunif(freq,20),scale=FALSE),T/freq)
d <- data.frame(t = 1:T)
b <- 0.5
c <- 0.1
d$y1 <-  b * d$t + seasonalindex + rexp(T,1/2)
d$y2 <- -b * d$t + seasonalindex + rexp(T,1/2)
 
d$y3 <- d$y1 + c * (d$t)^2 
d$y4 <- d$y1 - c * (d$t-25)^2 

d$y5 <- -d$y1 - c * (d$t)^2 
  
d$y6 <- -d$y1 + c * (d$t-25)^2 
  
m1 <- lm(y1~t,data=d)
library(broom)
g1 <- augment(m1,d) %>% ggplot(aes(x=t,y=y1)) +geom_line(alpha=0.5) + geom_point(alpha=0.5) +
  geom_line(aes(y=.fitted,x=t),col='red')  + theme(axis.title.x = element_blank(),axis.title.y = element_blank()) + ggtitle("Linear increase")

m2 <- lm(y2~t,data=d)
g2 <- augment(m2,d) %>% ggplot(aes(x=t,y=y2)) +geom_line(alpha=0.5) + geom_point(alpha=0.5) +
  geom_line(aes(y=.fitted,x=t),col='red')  + theme(axis.title.x = element_blank(),axis.title.y = element_blank())+ ggtitle("Linear decrease")


m3 <- lm(y3~t+I(t^2),data=d)
g3 <- augment(m3,d) %>% ggplot(aes(x=t,y=y3)) +geom_line(alpha=0.5) + geom_point(alpha=0.5) +
  geom_line(aes(y=.fitted,x=t),col='red')  + theme(axis.title.x = element_blank(),axis.title.y = element_blank()) + ggtitle("Accelerating increase")


m4 <- lm(y4~t+I(t^2),data=d)
g4 <- augment(m4,d) %>% ggplot(aes(x=t,y=y4)) +geom_line(alpha=0.5) + geom_point(alpha=0.5) +
  geom_line(aes(y=.fitted,x=t),col='red') + theme(axis.title.x = element_blank(),axis.title.y = element_blank()) + ggtitle("Decelerating increase")

m5 <- lm(y5~t+I(t^2),data=d)
g5 <- augment(m5,d) %>% ggplot(aes(x=t,y=y5)) +geom_line(alpha=0.5) + geom_point(alpha=0.5) +
  geom_line(aes(y=.fitted,x=t),col='red') + theme(axis.title.x = element_blank(),axis.title.y = element_blank()) + ggtitle("Accelerating decrease")

m6 <- lm(y6~t+I(t^2),data=d)
g6 <- augment(m6,d) %>% ggplot(aes(x=t,y=y6)) +geom_line(alpha=0.5) + geom_point(alpha=0.5) +
  geom_line(aes(y=.fitted,x=t),col='red') + theme(axis.title.x = element_blank(),axis.title.y = element_blank()) + ggtitle("Decelerating decrease")

(g1 + g3 + g4)/(g2 + g5 + g6)

```

---

## Trend

The long-term movement in a time series may occur due to many factors such as 

  -   changes in tastes and preferences of consumers, 
  
  -   technological changes, 
  
  -   demographic changes (changes in population and structure of population), 

  -   social and cultural changes
  
  -   etc.

---

## Trend

--

Analysis of trends in historical data helps evaluate past policies.

--

For example, the trend patterns of variables such as sales, revenues, and profits of a company for the past decade can tell us how successful the managerial policies of this period have been.

--

Knowing the trend of a time series is useful for making decisions about the future values of business and economic activities. 

--

An important question is how to extract the trend component from a time series.

- We can fit linear models with `lm()` to extract various trends from time series.
    
- We can fit more flexible 'smoother' lines using `geom_smooth()`.

---

### NZ resident population over time

```{r,echo=FALSE, fig.dim = c(10,5), out.width="100%",  message=FALSE}
gp1
```

---

### NZ resident population over time with 'smoother' line

```{r,echo=FALSE, fig.dim = c(10,5), out.width="100%", message=FALSE}
gp1 + geom_smooth(col=7) 
  
  
```

---

### NZ resident population over time with 'smoother' line and linear regression

```{r,echo=FALSE, fig.dim = c(10,5), out.width="100%", message=FALSE}
gp1 + geom_smooth(col=7) +
    geom_smooth(method='lm', col=2) + 
    xlab('Time') + ylab('NZ resident population')
```

---

### Total value of NZ housing stock

```{r,echo=FALSE, fig.dim = c(10,5), out.width="100%", message=FALSE}
gh1 <- housing %>% 
  ggplot() + 
  aes(x=date,y=total_value_of_housing_stock) + 
  geom_hline(yintercept=0) + 
  geom_point() + geom_line() +
  xlab('Time') + ylab('Total calue of NZ housing stock') +
  scale_y_continuous(labels = scales::dollar_format(prefix="$", scale=1/1e6, suffix = "T",accuracy=.5),
                     limits = c(-120000,1800000) )

gh1
```


---

### Total value of NZ housing stock with smoother

```{r,echo=FALSE, fig.dim = c(10,5), out.width="100%", message=FALSE}
gh2 <- gh1 + 
  geom_smooth(col=7) +
  annotate(geom="text", x = housing$date[40], y = 1.5e5, label="Smoother", col=7)
  
gh2
```

---

### Total value of NZ housing stock with smoother and linear trend

```{r,echo=FALSE, fig.dim = c(10,5), out.width="100%", message=FALSE}

gh3 <- gh2 + 
  geom_smooth(method='lm', formula = y ~ x, col=2) + 
  annotate(geom="text", x = housing$date[15], y = -60000, label="Linear", col=2) 
gh3 

```
  

---

### Total value of NZ housing stock with smoother and linear trend and quadratic trend

```{r,echo=FALSE, fig.dim = c(10,5), out.width="100%", message=FALSE}

gh3 + geom_smooth(method='lm', formula = y ~ x + I(x^2), col=4) +
  annotate(geom="text", x = housing$date[8], y = 270000, label="Quadratic", col=4)

```
  
  
---

class: middle, inverse

# Seasonal component

---

## Seasonal component 

--

Seasonal variations in a time series are defined as the movement that occurs in a time series periodically within a fixed time period. 

--

Many business activities, such as production and sales, exhibit seasonal patterns over different time periods (e.g. months or quarters) of a year. 

--

These seasonal variations may occur due to weather, holidays, or institutional factors. For example:

  - Sales of ski equipment are very high during winter and almost nonexistent during other seasons.
  - Sales at department stores around Christmas holidays are much higher than during other times of the year.
  - Power bills are expected to be higher during winter than in other seasons.

---

## Seasonal component 

--

Such fluctuations occur every year, although they can vary in magnitude. 

--

Nevertheless, with good data and appropriate statistical analysis, many seasonal variations in business and economic activities can be predicted.

--

An important question is how to extract the seasonal component from a time series.

  - We can use linear models with factors to extract seasonality from time series
    
  - Need to address trends and seasonality together!
    
  - The variability of seasonality may increase over time. This can be modelled by taking the log of $y$.
    
---

## Seasonality in NZ exports between 1990 to 2000

```{r,echo=FALSE, fig.dim = c(10,5), out.width="100%", message=FALSE}
library(lubridate)
overseastrade.ts <- overseastrade %>% 
  drop_na() %>% 
  mutate(time=1:n()) %>% 
  mutate(quarter=factor(quarter(date))) %>% 
  mutate(Quarter = recode_factor(quarter, '1'='Mar','2'='Jun','3'='Sep','4'='Dec')) %>%
  dplyr::filter(date < ymd('2000-01-01'))

rts.lm.q <- lm(export_values ~ time + quarter, data=overseastrade.ts)
gexp1 <- augment(rts.lm.q, overseastrade.ts) %>% 
  ggplot(aes(x=date, y=export_values)) + 
  geom_line(lwd=1, alpha=.5) + 
  geom_point(size =3, aes(col = Quarter)) + 
  scale_y_continuous(labels = scales::dollar_format(prefix="$", 
                                                    scale=1/1000, 
                                                    suffix = "B", 
                                                    accuracy=1),
                     limits = c(3.4e3,6.2e3)) + 
  xlab('Time') + ylab('Export value per quarter') + ggtitle('Goods sold by kiwis to the world') 

gexp1
```
    
---

## Seasonality in NZ exports between 1990 to 2000

```{r,echo=FALSE, fig.dim = c(10,5), out.width="100%", message=FALSE}
gexp1 + geom_line(aes(y=.fitted),col=1, lwd=1.2)+
  annotate(geom="text", x = housing$date[32], y = 6000, label="Model with trend and quarter", col=1)

```

---

class: middle, inverse

# Irregularity

---

## Irregular component

--

Irregular fluctuations are random or chance variations in a time series. 

--

They are the residuals after trend, *cyclical*, and seasonal components have been removed from the time series data.

--

Such movements may occur due to unpredictable events such as strikes, wars, earthquakes, hurricanes, floods, etc. 

--

These events can significantly affect the production of goods and services and other economic and business activities.

--

For example, floods usually cause a reduction in agricultural and industrial production. 

---

## Irregularity or residuals?

--

Due to the nature of irregular variations, they are impossible to predict.

--

They are similar to residuals after removing trend and seasonality from the time series regression!

--

However, we have observed that the residuals in time series regression look peculiar. 

--

They are not independent! 

--

Something evil lurks in the irregular component.

---

## Residuals of NZ housing

```{r,echo=FALSE,fig.dim = c(10,5), out.width="100%", message=FALSE}
housing.ts <- housing %>% mutate(time=1:n())
hpi.qm <- lm(total_value_of_housing_stock ~ time+I(time^2), data=housing.ts)

gh1 +
  geom_smooth(data = augment(hpi.qm, housing.ts),
              method='lm', formula = y ~ x + I(x^2), col=4) +
  # geom_smooth(alpha=0.2) + 
  geom_hline(yintercept=0) + 
  geom_point(data = augment(hpi.qm, housing.ts),
            aes(y=.resid), col=6, size=1.2) +
  geom_line(data = augment(hpi.qm, housing.ts),
            aes(y=.resid), col=6, lwd=1) +
  annotate(geom="text", x = housing$date[7], y = 270000, label="Quadratic fit", col=4) +
  annotate(geom="text", x = housing$date[70], y = 220000, label="Residuals", col=6)

```


---

class: middle, inverse

# Cycle

---


## Cyclical component

--

Usually a time series exhibits fluctuations around the trend line. 

--

That is, for some periods the actual time series curve is above the trend line and for other periods it is below the trend line. 

--

The non-periodic fluctuations in time series data around the trend line are referred to as the cyclical component or cyclical fluctuations. 

--

Note that the cyclical fluctuations do not last for any specific amount of time. They may last from more than one year to a period of 20 years.

--

Recession, depression, recovery and boom cause variations in time series
data, and these variations make up business cycles. 

--

These business cycles fluctuations are called the cyclical component or fluctuations.

---

## Cyclicity versus seasonality

--

A seasonal pattern exists when a series is influenced by seasonal factors
(e.g., the quarter of the year, the month, or day of the week). 

  - Seasonality is always of a fixed and known period no larger than one year.

--

A cyclic pattern exists when data exhibit rises and falls that are not of fixed period. 

  - The duration of these fluctuations is usually of at least 2 years.

--

The two are very different: 

  - If the fluctuations are not of fixed period then they are cyclic; 
    
  - if the period is unchanging and associated with some aspect of the calendar, then the pattern is seasonal.

---

## What can I do with the cyclic component?

--

This can be the component that has the most to offer in terms of understanding where the series may be headed in the short/medium term.

--

The length and amplitude of previous cycles may forecast the next turning point in the current cycle.

--

"**20/20 hindsight**" can often explain cyclic movements around the trend line in terms of past events or other variables

--

By looking at those variables or events in the present, we may be able to predict the likely future direction of the cycle movement.

--

**The Theory of Mean Reversion.**

---

## Business cycles

--

There are a variety of economic indicators, such as interest rates, unemployment rates, migration rates, etc.

--

It is possible that some of these, or an index summarising them, may be useful in predicting the cycle factor in a time series.

--

This could be done in regression analysis with the cycle factor (an estimate of the position in the cycle) as a dependent variable. 

--

Forecasting would involve predicting the rest of the cycle (not sure we're making progress here ...) using the index as the independent variable in a regression model.

--

In this course, we won't emphasize the cyclical component as we may lump it together with the residuals.

---

## Cycle of NZ housing?

```{r,echo=FALSE, fig.dim = c(10,5), out.width="100%", message=FALSE}
housing.ts <- housing %>% mutate(time=1:n())
hpi.qm <- lm(total_value_of_housing_stock ~ time+I(time^2), data=housing.ts)

gh1 +
  geom_smooth(data = augment(hpi.qm, housing.ts),
              method='lm', formula = y ~ x + I(x^2), col=4) +
  # geom_smooth(alpha=0.2) + 
  geom_hline(yintercept=0) + 
  geom_point(data = augment(hpi.qm, housing.ts),
            aes(y=.resid), col=6, size=1.2) +
  geom_line(data = augment(hpi.qm, housing.ts),
            aes(y=.resid), col=6, lwd=1) +
  annotate(geom="text", x = housing$date[7], y = 270000, label="Quadratic fit", col=4) +
  annotate(geom="text", x = housing$date[70], y = 220000, label="Residuals", col=6) +
  xlab('Time') + ylab('Total Value of Housing Stock') 
```



---

## Assemble or decompose? 

--

The components can be integrated via addition:     
$Y_t=\mbox{Trend}_t+\mbox{Cycle}_t+\mbox{Seasonality}_t+\mbox{Irregularity}_t$

--

The components can also be integrated via multiplication:     
$Y_t=\mbox{Trend}_t\times\mbox{Cycle}_t\times\mbox{Seasonality}_t\times\mbox{Irregularity}_t$

--

The above two are equivalent by log-transforming $y$.

--

Linear models are just one way to do time series decomposition. 

--

More will be covered in future statistics papers. 

---

## Summary

- Time series models and visualisation

- Four components in time series:
  - trend
  - seasonality
  - cyclicity
  - irregularity