---
title: 'Lecture 5'
subtitle: 'Reproducible analyses and data wrangling'
output:
  xaringan::moon_reader:
    css: [default, default-fonts, "custom.css"]
    nature:
      highlightStyle: tomorrow
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(palmerpenguins)
#data(package = 'palmerpenguins')
knitr::opts_chunk$set(echo = TRUE, comment = "")
knitr::opts_chunk$set(fig.dim=c(4.8, 4.5), fig.retina=2, out.width="100%")
```

class: middle, inverse

## Learning outcomes

- Reproducibility

- Data wrangling with dplyr

---

## Data today

We'll use the data from Horizons Regional council on river quality again today.

```{r, echo=FALSE, message=FALSE}
hrc <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/hrc/horizons_river_ecoli.csv")
hrc
```

These data are available from Land, Air, Water Aoteoroa, https://www.lawa.org.nz/download-data/

---

class: middle,inverse

# Reproducibility

---

## Reproducibility

A really important concept in science in general is of reproducibility.

We want to make sure that our findings are able to be replicated by:
 1. Ourselves using our data
 2. Others using our data
 3. Others using new data

In science, this is usually done by publishing our work with enough detail in the methods to allow replication by others.

But, often this is just replication in theory: In practice others will need much more than what most science teams provide in their publication.

---

## Open data

A big push in recent years by scientific bodies and publishers is to ensure that the data used in a piece of research is (as much as possible) published and freely available alongside the methods.

This helps replication, in that it's now possible for others to replicate using our data.

But, it's not everything! Often the methods used are not summarised in full detail, and typically there's a lot of reinventing the wheel that
needs to be done. e.g.

 - What data cleaning is required?
 
 - Were there missing data? How was it handled?
 
 - If some charts were produced, how were they generated? Was there manual manipulation (like adding labels/tidying plots by hand?)
 
 - What about statistical modelling? What was the exact formulation of the model? Are enough details given to allow replication?
 
---
 
## Principles of reproducible analyses

- Input data is documented and provided, and comes in whatever form it comes (i.e. we don't alter the data if it comes from another source).

- All analyses are scripted. i.e. run script(s) to produce all necessary tables/charts.

- No manual steps if at all possible.

- Ensure any scripting is useful for new data as much as possible.

https://github.com/jmarshallnz/dirichlet_island

https://royalsocietypublishing.org/doi/full/10.1098/rsif.2018.0534

---

class: middle, inverse

# Demo: NZ politics twitter sentiment by Kate Newton

https://www.rnz.co.nz/news/national/421955/bad-vibes-rating-political-scandals-by-twitter-toxicity

---

## Data wrangling

You'll have noticed in the last demo that it contained a bunch of code to wrangle data into a form that is usable for charting or tables.

Much of the data wrangling I do daily (and it is daily!) can be done using only a handful of functions, the majority taken from the package `dplyr`, with a few extras from other packages such as `tidyr`.

Essentially, the `dplyr` package provides a bunch of "verbs" (functions) for transforming data. It is similar to SQL if you have done any database work.

It is part of the `tidyverse` set of packages: https://www.tidyverse.org

---

class: middle, inverse

# `dplyr`

---

## `dplyr`

Just like `ggplot2` is a grammar of graphics, `dplyr` is a grammar of data manipulation.

It consists of a set of verbs that help us solve the most common data manipulation challenges:

- `select` to choose variables (columns) by name.
- `filter` to choose observations (rows) by value.
- `mutate` to add new variables based on existing variables.
- `summarise` to reduce multiple values down to a single summary.
- `arrange` changes the order of rows.

All of these combine with `group_by` to do all the above operation on subgroups of the data.

---

## `dplyr`

One of the nice things about `dplyr` is it abstracts the language of data processing across the implementation of where the data exist.

It works with `data.frame`'s loaded into memory, as well as remote databases (via SQL, or Apache Spark) using the same set of R commands.

The other nice thing is the consistency:
 - All verbs take in as their first argument a `data.frame`.
 - Then all other arguments describe what to do with the `data.frame`.
 - The result is then a new `data.frame`.

```{r eval=FALSE}
new_data <- do_something(old_data, other, stuff)
```
 
---

## Filtering rows with `filter`
 
Let's look again at the river data:
 
```{r}
hrc
```

---

## Filtering rows with `filter`
 
What if we just wanted the rows from 2017 onwards?
 
```{r}
filter(hrc, Date >= '2017-01-01')
```

---

## Filtering rows with `filter`
 
What if we wanted rows with large counts?

```{r}
filter(hrc, Value > 100000)
```

---

## Filtering rows with `filter`
 
What if we wanted really low counts before 2010?

```{r}
filter(hrc, Value < 100, Date <= '2010-12-31')
```

---

## Filtering rows with `filter`
 
What if we wanted the extreme low or high counts?

```{r}
filter(hrc, Value < 3 | Value > 100000)
```

---

## Logical expressions in R

The `filter` argument is a logical expression. i.e. it's something that evaluates to `TRUE` or `FALSE`.

For convenience, filter can take more than one logical expression, separated by a comma. In this case they're AND'd together (i.e. rows must satisfy both criteria), but anything within each expression is done first.

If we want to use OR rather than AND, we use the vertical bar `|` character.

i.e. `filter(hrc, Value < 3 | Value > 100000, SiteID == "00081")`

will return rows where the value is very small or very large from Site 00081.

---

## Logical expressions in R

Notice that for 'is equal to' we've been using `==`. The single `=` is for assignment of parameters instead. `dplyr` will warn you about this:

```{r, error=TRUE}
filter(hrc, Value = 4)
```

**This is rare** - usually error messages are rather more opaque!

Reading carefully, and searching the error on google can sometimes help.

---

## Logical expressions in R

Other operations are `!=` (not equal to) and `%in%` (in). e.g.

```{r}
filter(hrc, Value %in% c(3,5))
```

---

## Dealing with missing values

Notice the `Symbol` column is full of `NA` which means "missing". Maybe there are some that aren't missing? We might try:

```{r}
filter(hrc, Symbol != NA)
```

This seems like it might have worked?? It says there are no rows returned...

--

```{r}
filter(hrc, Symbol == NA)
```

Hmm, that doesn't seem right??

---

## Dealing with missing values

.left-code[
The key thing to remember is if a value is missing, then comparing it is senseless:

```{r}
2 == NA
NA == NA
```

Is 2 equal to something missing?

Maybe! We don't know! So the answer is missing.
]

--
.right-plot[
We use `is.na()` instead for this.

```{r}
is.na(2)
is.na(NA)
```
]

RStudio will remind you about this with a yellow warning in the margin, as it's a common problem.
---

## Dealing with missing values

Notice the `Symbol` column is full of `NA` which means "missing". Maybe there are some that aren't missing?

```{r}
filter(hrc, !is.na(Symbol))
```

Ok, it seems like we have a few (`r hrc |> filter(!is.na(Symbol)) |> nrow()`) numbers where the count actually means less than or greater than the value given, rather than equal to the value given.

---

## Arranging rows with `arrange`

Rather than pick rows out, we can sort the data instead.

```{r}
arrange(hrc, Value)
```

---

## Arranging rows with `arrange`

Selecting more than one variable to arrange on will first arrange on the first, then use later variables for ties:

```{r}
arrange(hrc, Value, Date)
```

---

## Arranging rows with `arrange`

To arrange descending rather than ascending, we can use `desc`

```{r}
arrange(hrc, Value, desc(Date))
```

Interestingly, no value of 1 after 2012... This suggests a change in the way things were measured.

---

## Combining `filter` and `arrange`

Let's look at post-2012 values:

```{r}
after2012 <- filter(hrc, Date >= '2013-01-01')
arrange(after2012, Value)
```

It seems some of the values recorded as 4 actually mean "less than 4".

---

## Combining `filter` and `arrange`

```{r}
large_values <- filter(hrc, Value > 50000)
arrange(large_values, Site, Date)
```

---

## A pipeline of verbs

Each of the verbs in `dplyr` take a dataset in as the first parameter, and return a data frame out again.

Often, an analyses will consist of a bunch of verbs called one after the other, with each one feeding into the next.

```{r, eval=FALSE}
large_values <- filter(hrc, Value > 50000)
arrange(large_values, Site, Date)
```

The temporary dataset `large_values` is really only used here to feed to the `arrange()` verb. We don't really need it.

We could do:

```{r, eval=FALSE}
arrange(filter(hrc, Value > 50000), Site, Date)
```

But you have to read that 'inside out'!

---

class: middle, inverse

# The pipe `|>`

---

## The pipe `|>`

The pipe operator, `|>` is specifically designed to improve this. It takes the item on the left of the pipe and places it
as the first argument of the function on the right.

```{r, eval=FALSE}
hrc |> filter(Value > 5000)
```

Is equivalent to:

```{r, eval=FALSE}
filter(hrc, Value > 5000)
```

But, the pipe version is readable as "Take the dataset `hrc` and filter it so that rows with Value > 5000 remain"

To enter a pipe in RStudio, you can use Ctrl-Shift-M (Command-Shift-M on a Mac).

---

## The pipe `|>`

This extends naturally:

```{r, eval=FALSE}
hrc |>
  filter(Value > 5000) |>
  arrange(Site, Date)
```

reads "Take the dataset `hrc`, filter it so that rows with Value > 5000 remain, then arrange by Site and Date"

When constructing pipelines like this, make sure the `|>` symbol is at the end of intermediate lines.

.left-code[
Use this:

```{r, eval=FALSE}
hrc |>
  filter(Value > 5000) |>
  arrange(Site, Date)
```
]

.right-plot[
Not this:

```{r, eval=FALSE}
hrc
 |> filter(Value > 5000)
 |> arrange(Site, Date)
```
]

---

## The magrittr pipe `%>%`

Prior to R version 4.1, the `|>` (base) pipe didn't exist. Instead, people used a pipe from the `magrittr` package which uses `%>%`.

Generally, these are basically the same. You will see `%>%` used 'in the wild' when googling - that's OK, it'll work just fine.

In RStudio, by default the keyboard combination Ctrl-Shift-M (Cmd-Shift-M on Mac) inserts the magrittr pipe `%>%`.

**We suggest you change this in RStudio**

See Tools->Global Options->Code and select the 'use native pipe operator, |>' option.

---

## Selecting and renaming columns

Just like we can select rows with `filter`, we can choose columns with `select`.

This is useful for getting rid of columns we don't need, for rearranging columns, and for changing names to
something more convenient.

```{r}
hrc |> select(Site, Date, Value)
```

---

## Selecting columns with `select`

We can select a range of columns with a colon `:`

```{r}
hrc |> select(Council, Site:RawValue)
```

---

## Selecting columns with `select`

We can select by column number instead of column name if you like:

```{r}
hrc |> select(Council, 3:6)
```

Generally it's better to use names though, as the order of columns might be different to what you expect.

---

## Selecting columns with `select`

The `everything()` helper function is useful if you want one column first, then everything else:

```{r}
hrc |> select(Symbol, Value, everything())
```

---

## Selecting columns with `select`

The `starts_with()` helper is useful if you have a bunch of columns that start similarly:

```{r}
hrc |> select(Date, Value, starts_with("Site"))
```

---

## Selecting columns with `select`

You can rename a column just by changing it's name in the `select`:

```{r}
hrc |> select(Site, Date, EColi = Value)
```

---

## Dropping columns

To drop a column use `-` in front of the name:

```{r}
hrc |> select(-Indicator, -RawValue, -Symbol)
```

---

## Renaming columns with `rename`

If you want to rename a column but don't want to specify all the other columns, then `rename()` does this:

```{r}
hrc |> rename(Name = Site)
```

---

## Adding new columns: `mutate`

We can add new columns based on existing ones with `mutate`:

```{r}
hrc |> select(SiteID, Date, Value) |>
  mutate(Log10Value = log10(Value))
```

---

## Adding new columns: `mutate`

We can add new columns based on existing ones with `mutate`:

```{r, message=FALSE}
library(lubridate)
hrc |> select(SiteID, Date, Value) |>
  mutate(Year = year(Date), Month = month(Date, label=TRUE))
```

---

.left-code[
## Seasonality?
```{r, label='hrc_season', eval=FALSE}
hrc |>
  mutate(
    Month = month(Date, label=TRUE)
    ) |>
  ggplot() +
  geom_boxplot(
    aes(
      x = Month,
      y = Value)
    ) +
  scale_y_log10(
    labels = scales::label_comma()
  )
```

We can pipe the result of wrangling into
`ggplot()` as it takes a data.frame as first
argument.

Remember to switch from `|>` to `+` !!
]

.right-plot[
```{r, ref.label='hrc_season', echo=FALSE, eval=TRUE}
```
]

---

## Adding new columns: `mutate`

We can change existing columns by overwriting them:

```{r}
library(lubridate)
hrc |> select(SiteID, Date, Value) |>
  mutate(SiteID = as.numeric(SiteID))
```

---

## Adding new columns: `mutate`

We can change existing columns by overwriting them:

```{r}
hrc |> select(SiteID, Site, Date, Value) |>
  mutate(SiteID = as.numeric(SiteID),
         Site = str_to_title(Site))
```

---

## How I extracted value, symbol

.left-code-wide[
```{r, label="hrc_extract", eval=FALSE}
hrc |> select(SiteID, RawValue) |>
  mutate(
    Symbol = str_extract(RawValue, '[<>]'),
    Value = str_extract(RawValue, '[0-9]+'),
    Value = as.numeric(Value)
    ) |>
  arrange(Symbol)
```

The patterns `[<>]`, `[0-9]+` are **regular expressions**.

They're a way of specifying how to match parts of text.

Don't worry about these - you won't have to come up with them yourselves!
]

.right-out-narrow[
```{r, ref.label="hrc_extract", echo=FALSE, eval=TRUE}
```
]

---

## To summarise

Today we've learnt about:

- `filter()` to pick rows based on value.

- `arrange()` to change the order of rows based on value.

- `select()` to pick columns based on column name.

- `rename()` to rename columns while keeping others.

- `mutate()` to add new columns based on existing ones.

--

The real power in `dplyr` comes from the ones we'll look at next time:

- `summarise()` to reduce all rows down to a summary statistic.

- `group_by()` to do all of the above operations within subgroups of the data all at once.
