---
title: 'Lecture 2'
subtitle: 'Charting with R using `ggplot2`'
output:
  xaringan::moon_reader:
    css: [default, default-fonts, "custom.css"]
    nature:
      highlightStyle: tomorrow
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
      slideNumberFormat: |
        <div class="progress-bar-container">
          <div class="progress-bar" style="width: calc(%current% / %total% * 100%);">
          </div>
        </div>
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(palmerpenguins)
penguins = na.omit(penguins)
#data(package = 'palmerpenguins')
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.dim=c(4.8, 4.5), fig.retina=2, out.width="100%")
```

## Learning outcomes

- Mapping aesthetics versus setting values

- Titles and axis labels

- Scales and colours

---

## Data today

We'll use the penguin data again, but also some from Horizons Regional council on river quality. In particular, counts of *E. coli* as an indicator of faecal contamination.

```{r, message=FALSE}
hrc <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/hrc/horizons_river_ecoli.csv")
hrc
```

These data are available from Land, Air, Water Aotearoa, https://www.lawa.org.nz/download-data/

---

## Data today

We have *E. coli* counts from `r hrc |> count(Site) |> nrow()` sites throughout the horizons regional council.

```{r}
count(hrc, Site)
```

---

## Data today

And the counts can be summarised as

```{r}
summarise(hrc, min(Value), max(Value), median(Value))
```

You can see that we have very low counts (1) and very high counts (over 200,000), with half the counts being less than 155.

---

class: middle, inverse

# Remember the recipe!

---

## Remember the Recipe!

- Tell R we want a plot with `ggplot()`.

- Tell `ggplot` what data we are plotting with `data=penguins`.

- Choose what type of chart we want by adding a layer (geometry), e.g. `geom_point()`.

- Tell the geometry how to map from visual content/features to data with `mapping = aes()`.

---

.left-code[
## Remember the Recipe!
```{r peng2, eval=FALSE}
ggplot(data = penguins) + #<<
  geom_point(
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g
      )
    )
```
Tell `ggplot` what data we want to plot.
]

.right-plot[
```{r ref.label="peng2", echo=FALSE}
```
]

---

.left-code[
## Remember the Recipe!
```{r peng3, eval=FALSE}
ggplot(data = penguins) +
  geom_point( #<<
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g
      )
    ) #<<
```
Choose the type of chart we want.
]

.right-plot[
```{r ref.label="peng3", echo=FALSE}
```
]

---

.left-code[
## Remember the Recipe!
```{r peng4, eval=FALSE}
ggplot(data = penguins) +
  geom_point(
    mapping = aes( #<<
      x = flipper_length_mm, #<<
      y = body_mass_g #<<
      ) #<<
    )
```

Map plot features to data features.
]

.right-plot[
```{r ref.label="peng4", echo=FALSE}
```
]

---

class: middle, inverse

# Aesthetics

## Setting or mapping features of the plot

---

## Aesthetics

The visual content or features of the plot are **aesthetics**. Things like:

- position of points (`x`, `y`)

- height of bars (`y`)

- colour of points (`colour`, `alpha`)

- colour inside regions (`fill`)

- point type and size (`shape`, `size`)

All of these can either be **set to a single value**, or allowed to **change by mapping them to data**.

---

.left-code[
## Setting a colour
```{r peng5, eval=FALSE}
ggplot(data = penguins) +
  geom_point(
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g
      ),
    colour = 'purple' #<<
    )
```

To **set** to a single value, place inside the `geom_*()` function.
]

.right-plot[
```{r, ref.label="peng5", echo=FALSE}
```
]

---

.left-code[
## Setting versus Mapping
```{r peng6, eval=FALSE}
ggplot(data = penguins) +
  geom_point(
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g,
      colour = 'purple' #<<
      )
    )
```
To **set** to a single value, place inside the `geom_*()` function.

But not inside the `aes()` function!

This is like adding a new column, and setting every entry to `purple`, so you have a grouping variable with one group `purple`.
]

.right-plot[
```{r, ref.label="peng6", echo=FALSE}
```
]

---

.left-code[
## Mapping colour to data
```{r peng7, eval=FALSE}
ggplot(data = penguins) +
  geom_point(
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g,
      colour = species #<<
      )
    )
```

To **map to a column of data** place inside the `aes()` function.
]

.right-plot[
```{r, ref.label="peng7", echo=FALSE}
```
]

---

## Mapping aesthetics

Aesthetics are mapped to data via the `aes()` function passed to the `mapping` parameter of each geometry layer.

What this does is define a map between features of the plot (`x`, `y`, `colour`) and features of our data (columns).

When this is done, `ggplot` automatically adds a scale or guide so you know which styling maps
to which parts of your data.

e.g. the `x` and `y` aesthetics automatically have an axis scale. The `colour` aesthetic will give a legend or colour scale.

**Any time you want a feature of the plot to change based on data, it needs to be in `aes()`**

---

## Different species

```{r}
aes(x = flipper_length_mm, 
    y = body_mass_g,
    colour = species)
```

---

.left-code[
## Different species
```{r peng8, eval=FALSE}
ggplot(data = penguins) +
  geom_point(
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g,
      colour = species #<<
      )
    )
```
`colour` changes with data (mapping).
]

.right-plot[
```{r, ref.label="peng8", echo=FALSE}
```
]

---

.left-code[
## Different species
```{r peng9, eval=FALSE}
ggplot(data = penguins) +
  geom_point(
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g,
      size = species #<<
      ),
    alpha = 0.5 #<<
    )
```
`size` changes with data (mapping).

`alpha` is fixed for all points (setting).
]

.right-plot[
```{r, ref.label="peng9", echo=FALSE}
```
]

---

.left-code[
## Different species
```{r peng10, eval=FALSE}
ggplot(data = penguins) +
  geom_point(
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g,
      shape = species #<<
      ),
    size = 2 #<<
    )
```
`shape` changes with data (mapping).

`size` is fixed for all points (setting).
]

.right-plot[
```{r, ref.label="peng10", echo=FALSE}
```
]

---

class: middle, inverse

# Geometry layers

---

## Other layers/geometries

Instead of the `geom_point` layer, we could use something else.

Last time I counted there were 50 different geometry layers to choose from.

And there are add-on packages with more!

---

.left-code[
## `geom_point()`
```{r peng11, eval=FALSE}
ggplot(data = penguins) +
  geom_point( #<<
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g
      )
    )
```
]

.right-plot[
```{r, ref.label="peng11", echo=FALSE}
```
]

---

.left-code[
## `geom_smooth()`
```{r peng12, eval=FALSE}
ggplot(data = penguins) +
  geom_smooth( #<<
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g
      )
    )
```
]

.right-plot[
```{r, echo=FALSE, message=FALSE}
ggplot(data = penguins) +
  geom_point(mapping = aes(x=flipper_length_mm, y=body_mass_g), alpha=0) +
  geom_smooth(
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g
      )
    )
```
]

---

.left-code[
## Adding layers
```{r peng13, eval=FALSE}
ggplot(data = penguins) +
  geom_point(
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g
      )
    ) +
  geom_smooth( #<<
    mapping = aes( #<<
      x = flipper_length_mm, #<<
      y = body_mass_g #<<
      ) #<<
    ) #<<
```
]

.right-plot[
```{r, ref.label="peng13", echo=FALSE, message=FALSE}
```
]

---

.left-code[
## Too much typing!
```{r peng14, eval=FALSE}
ggplot(data = penguins) +
  geom_point(
    mapping = aes( #<<
      x = flipper_length_mm, #<<
      y = body_mass_g #<<
      ) #<<
    ) +
  geom_smooth(
    mapping = aes( #<<
      x = flipper_length_mm, #<<
      y = body_mass_g #<<
      ) #<<
    )
```
]

.right-plot[
```{r, ref.label="peng14", echo=FALSE, message=FALSE}
```
]

---


.left-code[
## Inheriting aesthetics
```{r peng15, eval=FALSE}
ggplot(data = penguins,
       mapping = aes( #<<
         x = flipper_length_mm, #<<
         y = body_mass_g #<<
         ) #<<
       ) +
  geom_point() +
  geom_smooth()
```
Any mapping defined in the `ggplot()` function are inherited by other layers.
]

.right-plot[
```{r, ref.label="peng15", echo=FALSE, message=FALSE}
```
]

---

.left-code[
## Unique aesthetics
```{r peng16, eval=FALSE}
ggplot(data = penguins,
       mapping = aes(
         x = flipper_length_mm,
         y = body_mass_g
         )
       ) +
  geom_point(
    mapping = aes( #<<
      colour = species #<<
      ) #<<
    ) +
  geom_smooth()
```
You can override or add new aesthetics in each item.
]

.right-plot[
```{r, ref.label="peng16", echo=FALSE, message=FALSE}
```
]


---

class: middle, inverse

# Naming things

---

## Naming things: code style

R code uses a lot of **functions**. The basic syntax for a function is:

```{r, eval=FALSE}
function_name(param1, param2, param3)
```

Functions can take a bunch of **parameters**. These are matched by name (if named), else by position.

The code for our plot contains 3 functions:
 - `ggplot(data = penguins)`
 - `geom_point(mapping = aes(...))`
 - `aes(x = flipper_length_mm, y = body_mass_g)`

Note in each case we've named the parameters (i.e. `data = penguins`, `x = flipper_length_mm`). This is generally good practice. If unnamed you're relying on order, so this works too:

```{r, eval=FALSE}
ggplot(penguins) + geom_point(aes(flipper_length_mm, body_mass_g))
```

---

## Naming things: titles and labels

Charts should, as much as possible, be self-contained. They should tell the story without the viewer/reader having to figure out too much what each bit of the plot represents.

- Concise, accurate axis labels, including units where relevant.

- Concise, clear legend names and legend items.

- A title (and possibly subtitle) that tells the story.

- Possibly further captioning that helps (e.g. data sources, authorship).

---

## Axis labels

Use `labs()` to change axis labels (or any labels associated with an aesthetic). 

```{r, eval=FALSE}
labs(x = "The x axis label I want to use",
     y = "The y axis label",
     colour = "The legend title for colour")
```

---

.left-code[
## Axis labels
```{r peng17, eval=FALSE}
ggplot(data = penguins,
       mapping = aes(
         x = flipper_length_mm,
         y = body_mass_g
         )
       ) +
  geom_point(
    mapping = aes(
      colour = species
      )
    ) +
  geom_smooth() +
  labs(x = "Flipper length (mm)", #<<
       y = "Body mass (g)") #<<
```
]

.right-plot[
```{r, ref.label="peng17", echo=FALSE, message=FALSE}
```
]

---

## Titles and captions

Titles should inform the reader what the chart is showing.

**Ideally the title should help tell the story.**

This is particularly important for charts shared with others (e.g. on social media).

Anything you can add to the chart that helps tell the story means that those who see the chart immediately know what it is showing them.

You can set a chart title (and subtitles or captions) using

```{r, eval=FALSE}
labs(title = "A useful title",
     subtitle = "Shown smaller, under the title",
     caption = "Shown in the bottom-right, useful for data source")
```

BUT! Be aware of editorialising too much. We've all read articles where the headlines mislead!

---

.left-code[
## Titles tell the story
```{r peng18, eval=FALSE}
ggplot(data = penguins,
       mapping = aes(
         x = flipper_length_mm,
         fill = species
         )
       ) +
  geom_density(alpha = 0.5) +
  labs(x = "Flipper length (mm)",
       title = "Distribution of penguin flipper lengths")
```
This is a bad title. We can tell it's a distribution already!

"Penguin" is the only useful bit.
]

.right-plot[
```{r, ref.label="peng18", echo=FALSE, message=FALSE}
```
]

---

.left-code[
## Titles tell the story
```{r peng19, eval=FALSE}
ggplot(data = penguins,
       mapping = aes(
         x = flipper_length_mm,
         fill = species
         )
       ) +
  geom_density(alpha = 0.5) +
  labs(x = "Flipper length (mm)",
       title = "Penguin flipper length differs by species")
```
This is better.

But surely it is obvious that flipper length would differ by species!
]

.right-plot[
```{r, ref.label="peng19", echo=FALSE, message=FALSE}
```
]

---

.left-code[
## Titles tell the story
```{r peng20, eval=FALSE}
ggplot(data = penguins,
       mapping = aes(
         x = flipper_length_mm,
         fill = species
         )
       ) +
  geom_density(alpha = 0.5) +
  labs(x = "Flipper length (mm)",
       title = "Gentoo penguins have longer flippers")
```
This is even better.

It's making the story clear.
]

.right-plot[
```{r, ref.label="peng20", echo=FALSE, message=FALSE}
```
]

---

```{r, echo=FALSE, message=FALSE, fig.dim=c(6,3.5), fig.retina=3}
border_plot <- read_csv(here::here("data/covid_arrivals/border_plot.csv"))
ggplot(border_plot) +
  aes(x=Date, y=fit) +
  geom_col(aes(x=Date, y=cases_per_arrival), fill='red') +
  geom_ribbon(aes(ymin=lwr, ymax=upr), alpha=0.3) +
  geom_line() +
  ylab("Detected COVID-19 positives (%)") +
  xlab("") +
  theme_minimal(base_size=10) +
  ggtitle("The Mystery of the missing May: COVID-19 cases at the border",
          subtitle = "Bars are cases per person, line and band is smoothed trend with uncertainty (GAM)")
```

---

class: middle, inverse

# Scales and colours

---

## Scales

Each aesthetic has an associated scale. `x` and `y` have axes, `colour` has a guide. These scales can involve a transformation (e.g. a log scale) or limits, or we can choose what colours to associate with each group.

We do this through the `scale_` functions. e.g.
  - `scale_x_log10()` to switch to a log scale,
  - `scale_y_continuous` to define limits on a numeric y-axis, 
  - `scale_colour_manual` for specifying colours manually.

And lots more for specific circumstances.

---

.left-code[
## Changing axis scales
```{r peng21, eval=FALSE}
ggplot(data = penguins) +
  geom_point(
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g
      )
    ) +
  scale_x_continuous( #<<
    limits = c(100, 300) #<<
    ) #<<
```
]

.right-plot[
```{r, ref.label="peng21", echo=FALSE, message=FALSE}
```
]

---

.left-code[
## Changing axis scales
```{r hrc1, eval=FALSE}
ggplot(data = hrc) +
  geom_point(
    mapping = aes(
      x = Date,
      y = Value
      ),
    alpha=0.5
  )
```
]

.right-plot[
```{r, ref.label="hrc1", echo=FALSE, message=FALSE}
```
]

---

.left-code[
## Changing axis scales
```{r hrc2, eval=FALSE}
ggplot(data = hrc) +
  geom_point(
    mapping = aes(
      x = Date,
      y = Value
      ),
    alpha=0.5
  ) +
  scale_y_log10() #<<
```
]

.right-plot[
```{r, ref.label="hrc2", echo=FALSE, message=FALSE}
```
]


---

.left-code[
## Changing axis scales
```{r hrc3, eval=FALSE}
ggplot(data = hrc,
       mapping = aes(
         x = Date,
         y = Value
       )
     ) +
  geom_point(alpha=0.1) +
  geom_smooth() +
  scale_y_log10()
```
]

.right-plot[
```{r, ref.label="hrc3", echo=FALSE, message=FALSE}
```
]

---

.left-code[
## Changing colours
```{r peng_col1, eval=FALSE}
ggplot(data = penguins) +
  geom_point(
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g,
      col = species
      )
    )
```
]

.right-plot[
```{r, ref.label="peng_col1", echo=FALSE, message=FALSE}
```
]

---

.left-code[
## Changing colours
```{r peng_col2, eval=FALSE}
ggplot(data = penguins) +
  geom_point(
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g,
      col = species
      )
    ) +
  scale_colour_manual( #<<
    values = c( #<<
      Adelie = "darkgreen", #<<
      Chinstrap = "maroon", #<<
      Gentoo = "orange" #<<
    ) #<<
  ) #<<
```
]

.right-plot[
```{r, ref.label="peng_col2", echo=FALSE, message=FALSE}
```
]

---

## Changing colours

The `c()` function is short for 'combine'. In this case we're combining 3 colours together, naming them after each species.

```{r, eval=FALSE}
c(Adelie = "darkgreen", Chinstrap = "maroon", Gentoo = "orange")
```

---

.left-code[
## Changing colours
```{r peng_col2b, eval=FALSE}
peng_colour = c( #<<
      Adelie = "darkgreen", #<<
      Chinstrap = "maroon", #<<
      Gentoo = "orange" #<<
    ) #<<
ggplot(data = penguins) +
  geom_point(
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g,
      col = species
      )
    ) +
  scale_colour_manual( #<<
    values = peng_colour #<<
  ) #<<
```
]

.right-plot[
```{r, ref.label="peng_col2b", echo=FALSE, message=FALSE}
```
]

---

.left-code[
## Changing colours: palettes
```{r peng_col3, eval=FALSE}
ggplot(data = penguins) +
  geom_point(
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g,
      col = species
      )
    ) +
  scale_colour_viridis_d() #<<
```
]

.right-plot[
```{r, ref.label="peng_col3", echo=FALSE, message=FALSE}
```
]

---

.left-code[
## Changing colours: palettes
```{r peng_col4, eval=FALSE}
ggplot(data = penguins) +
  geom_point(
    mapping = aes(
      x = flipper_length_mm,
      y = body_mass_g,
      col = species
      )
    ) +
  scale_colour_brewer( #<<
    palette = "Dark2" #<<
    ) #<<
```
]

.right-plot[
```{r, ref.label="peng_col4", echo=FALSE, message=FALSE}
```
]
