---
title: "Workshop A08: Finding the MP for Wairarapa's Ute"
output: html_document
---

## Introduction

Today we'll be looking at a subset from the NZTA fleet data. This holds registration information on all vehicles in New Zealand.

We have a subset of it which is loaded below. If you're on a slower internet connect, you might want to download the data from the URL and load it locally.

```{r message=FALSE}
library(tidyverse)

fleet <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/fleet.csv.gz")
#roll <- read_csv("fleet.csv.gz") # for loading from a local copy

fleet
```

## Reproducing the plot

The following code block filters the `fleet` data down to just cars and vans:

```{r}
cars_vans <- fleet |> filter(VEHICLE_TYPE == "PASSENGER CAR/VAN")
```

### Try yourself

1. Summarise the data by counting the number of cars/vans of each `BASIC_COLOUR` for each `VEHICLE_YEAR` and then convert these to proportions by `VEHICLE_YEAR`. *Hint: `group_by(VEHICLE_YEAR)` followed by a `count` and then `mutate` is what you need.*

```{r}
# Option 1: Group by both year and colour then summarise to compute number of rows with n()
cars_vans |>
  group_by(VEHICLE_YEAR, BASIC_COLOUR) |>
  summarise(n = n())

# Option 2: Group by year, then use count() to count number of rows of each colour.
cars_vans |>
  group_by(VEHICLE_YEAR) |>
  count(BASIC_COLOUR)

# Compute the proportion within year:
prop_colour_by_year = cars_vans |>
  group_by(VEHICLE_YEAR) |>
  count(BASIC_COLOUR) |>
  mutate(total = sum(n)) |>
  mutate(prop = n / total)

prop_colour_by_year
```

2. Get the basic plot up and running by plotting proportion on the y-axis and `VEHICLE_YEAR` on the x-axis, facetting by `BASIC_COLOUR`.

```{r}
ggplot(data = prop_colour_by_year) +
  geom_line(mapping = aes(x=VEHICLE_YEAR, y=prop)) +
  facet_wrap(vars(BASIC_COLOUR))
```

3. Fine tune the plot with labelling.

```{r}
ggplot(data = prop_colour_by_year) +
  geom_line(mapping = aes(x=VEHICLE_YEAR, y=prop)) +
  facet_wrap(vars(BASIC_COLOUR)) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(y="Percentage of cars and vans",
       title="The rise of orange and demise of brown and gold",
       subtitle="Car and van colours in New Zealand",
       x=NULL) +
  theme_bw()
```

## Are Ford Rangers a rural phenomenon?

The following code pulls out utes (utility vehicles) from 2016 onwards, and classifies them via a `Type` variable into `Ranger` or `Other`:

```{r}
utes <- fleet |> filter(BODY_TYPE == "UTILITY") |>
  filter(VEHICLE_YEAR > 2015) |>
  mutate(Type = ifelse(MAKE == "FORD" & MODEL == "RANGER", "Ranger", "Other"))
```

### Try yourself

1. Calculate the proportion of utes that are Ford Rangers for each TLA. *Hint: You want to `group_by(TLA)` and then `count`, then compute the proportion, as was done for colours above.*

```{r}
prop_rangers <- utes |>
  group_by(TLA) |>
  count(Type) |>
  mutate(total = sum(n)) |>
  mutate(prop = n/total)

prop_rangers
```

2. Using only the proportions that correspond to `Ranger` (rather than `Other`) utes, produce a bar chart for the proportion of Ford Rangers across each `TLA`. You might want to use `fct_reorder` to reorder the `TLA` by the proportion variable so the barchart is ordered. Remember to use `geom_col` for where you have already computed the bar lengths.

```{r, fig.height=10, fig.width=7}
prop_rangers |>
  filter(Type == "Ranger") |>
  ungroup() |>
  mutate(TLA = str_to_title(TLA)) |>
  mutate(TLA = fct_reorder(TLA, prop)) |>
  ggplot() +
  geom_col(mapping=aes(y=TLA, x=prop)) +
  scale_x_continuous(labels = scales::label_percent(), expand=c(0,0)) +
  labs(x = "Percentage of utes that are Rangers", 
       y = NULL,
       title = "Proportion of utes that are Ford Rangers",
       subtitle = "Some urban areas have a large proportion of utes that are Rangers") +
  theme_minimal()
```

3. What is your conclusion? Are Ford Rangers a rural thing?

Perhaps Rangers may be a urban thing as well as a rural thing.


## Finding the MP for Wairarapa's Ute

In this question your goal is to find the row in the dataset that contains the Ute owned by the Member of Parliament for the Wairarapa electorate. Use all the `fleet` data for this question.

### Try yourself:

1. Use Google to find out who the Member of Parliament for Wairarapa is, in case you don't know.

Kieran McAnulty is the MP for Wairarapa.

2. Use Google image search to find a picture of their ute. You want one where you can see the number plate (registration).

Number plate is GCU805

3. Lookup the registration on https://www.rightcar.govt.nz/ to find the brand, colour and year of the vehicle.

MAZDA B2500 1997 RED DBL CAB P/UP 2499, 64kW, 5 Gear manual, Diesel

4. Based on the `VEHICLE_YEAR`, `MAKE`, `BASIC_COLOUR` and that the `BODY_TYPE` is `UTILITY` and that it is registered in the `MASTERTON DISTRICT` (this is the `TLA`), find the row corresponding to this vehicle.

```{r}
fleet |>
  filter(VEHICLE_YEAR == 1997,
         MAKE == "MAZDA",
         BASIC_COLOUR == "RED",
         BODY_TYPE == "UTILITY",
         MODEL == "B2500",
         TLA == "MASTERTON DISTRICT",
         NUMBER_OF_SEATS == 5)
```

Done! :)