```{r data generation, echo = FALSE, results = "hide"}
x<-round(rcauchy(21),2)
answer<- length(boxplot(x, plot=F)$out)
other.answers<-sample(c(0:(max(answer,3)+2))[-(answer+1)],4)

data <- data.frame(x=x)
write.csv(data, "dataset.csv", row.names = FALSE)

questions <- character(5)
questions[1] <- answer
questions[2] <- other.answers[1]
questions[3] <- other.answers[2]
questions[4] <- other.answers[3]
questions[5] <- other.answers[4]

solutions <- c(TRUE,FALSE,FALSE,FALSE,FALSE)
explanations <- character(5)
explanations[1] <- ""
explanations[2] <- ""
explanations[3] <- ""
explanations[4] <- ""
explanations[5] <- ""

## permute order of solutions/questions
Order <- sample(1:5)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

Download the data file [dataset.csv](dataset.csv) by right clicking and 'Save link as', then load it into RStudio using `read_csv()` or the Import Dataset button.

Use `ggplot2` to make a boxplot of the `x` column. How many extreme values are shown using the default settings?

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

```{r boxplot_hist, message = FALSE, results = "hide", fig.height = 2, fig.width = 5, fig.path = "", fig.cap = ""}
library(tidyverse)
dataset <- read_csv("dataset.csv")
ggplot(dataset) + geom_boxplot(aes(x=x))
```

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================

extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice
