```{r, data_generation, echo=FALSE, results="hide"}
questions <- character(5)
questions[1] <- "Because both variables are discrete, with a small number of values, the points have plotted on top of each other."
questions[2] <- "I should have put the 'aes' part inside 'ggplot()'"
questions[3] <- "The only explanation is a data entry error."
questions[4] <-  "The `geom_point()` function can only plot up to 50 data points."
questions[5] <- "The variables are categorical so a bar chart should be used."

solutions <- c(TRUE,FALSE,FALSE,FALSE,FALSE)
explanations <- character(5)
explanations[1] <- "With a small number of discrete values there will be overplotting. Adding some jitter might help."
explanations[2] <-  "The `aes` mapping can go either in `ggplot()` or in the `geom_*()` function."
explanations[3] <- "Data entry error is unlikely to be the case 490 times!"
explanations[4] <- "The `geom_point()` function can take as many data points as you like."
explanations[5] <- "The variables could be interpreted as categorical as the counts are low, so a barchart is an option (e.g. number of households by number of children, coloured by number of cars). However, this is not the problem with the scatter plot."
## permute order of solutions/questions
Order <- sample(1:5)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

I take a sample of 500 households, observe the number of children and the number of cars in each household, and make a scatter plot, using the command `ggplot(data) + geom_point(aes(x=cars, y=children))`.  My scatter plot appears to have only 10 points on it.  What has gone wrong?

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================
extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple choice