```{r data generation, echo = FALSE, results = "hide"}
library(dplyr)
library(readr)
library(exams)

N<- sample(c(30,40,50),1)
X<-exp(runif(N)*10)+10
X<-round(X)

alpha <- data.frame(times=X)
write.csv(alpha, "alpha.csv", row.names = FALSE)

questions <- character(5)
questions[1] <- paste(round(quantile(X,c(0.25, .5, .75)),2), collapse=', ')
questions[2] <-  paste(round(quantile(X, c(0.15, .5, .85)),2), collapse=', ')
questions[3] <-  paste(round(quantile(X, c(0.40, 0.5, .6),2)), collapse=', ')
questions[4] <-  paste(round(mean(X)+c(-1*sd(X),0,sd(X)),0), collapse=', ')
questions[5] <-  paste(round(mean(X)+c(-2*sd(X),0, 2*sd(X)),0), collapse=', ')

solutions <- c(TRUE,FALSE,FALSE,FALSE,FALSE)
explanations <- character(5)
explanations[1] <- ""
explanations[2] <- ""
explanations[3] <- ""
explanations[4] <- ""
explanations[5] <- ""

## permute order of solutions/questions
Order <- sample(1:5)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

Download the data file [alpha.csv](alpha.csv) of the amount of time in minutes needed to resolve a sample of problems arising in a test of new software by right clicking and 'Save link as', then load it into RStudio using `read_csv()` or the Import Dataset button.

Use `dplyr` to find the lower quartile, median, and upper quartile of the distribution of times.

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

```{r air_iqr, echo = TRUE, message=FALSE, comment=""}
library(tidyverse)
alpha = read_csv("alpha.csv")
alpha |> summarise(quantile(times, probs=c(0.25, 0.5, 0.75)))
```

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================

extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice
