```{r data generation, echo = FALSE, results = "hide"}
library(dplyr)
library(readr)
library(exams)

odis <- read_csv("https://www.massey.ac.nz/~jcmarsha/161122/data/ODIs.csv")

# choose the variable(s)
selected_var <- "Ground"

possible_answers <- NULL
while (TRUE) {
  # OK, now for each one, find the one we want to filter to
  options <- odis |> pull(selected_var) |> unique()

  selected <- sample(options, 1)

  # check we have enough of what we want
  possible_answers <- odis |> filter(.data[[selected_var]] == selected) |>
    group_by(Team) |> summarise(Score = max(Score)) |>
    arrange(desc(Score))

  # restrict to 5
  if (nrow(possible_answers) >= 5) {
    rows <- c(1, sample(2:nrow(possible_answers), 4))
    possible_answers <- possible_answers |> slice(rows) |>
      arrange(desc(Score))
    # check the first two
    score_values <- possible_answers |> pull(Score)
    if (score_values[1] >= score_values[2]) {
      break
    }
  }
}

questions <- possible_answers$Team
solutions <- c(TRUE,rep(FALSE, length(questions)-1))

explanations <- character(length(solutions))

## permute order of solutions/questions
Order <- sample(seq_along(questions))
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

A dataset of one day cricket match results from 1971 to 2019 may be downloaded into RStudio using the following code:

```{r, eval=FALSE}
library(tidyverse)
odis <- read_csv("https://www.massey.ac.nz/~jcmarsha/161122/data/ODIs.csv")
```

Using these data, the `Team` that has the highest `Score` at the `Ground` "`r selected`" is:

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

```{r, echo=FALSE, message=FALSE, warning=FALSE, comment = "", results='asis'}
src <- knitr::knit_expand(text=c('```{r, message=FALSE, warning=FALSE}',
                   'library(tidyverse)',
                   'odis <- read_csv("https://www.massey.ac.nz/~jcmarsha/161122/data/ODIs.csv")',
                   'odis |>',
                   '  filter({{selected_var}} == "{{selected}}") |>',
                   '  arrange(desc(Score))',
                   '```'))
res = knitr::knit_child(text = src, options = list(comment = ""), quiet=TRUE)
cat(res)
```

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================

extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice
