```{r data generation, echo = FALSE, results = "hide"}
library(dplyr)
library(readr)
library(exams)
N=13+sample(10,1)
Min=20+sample(5,1)
Scalar=5+sample(4,1)
X=round(Min + Scalar*runif(N), 2)

air <- data.frame(CO=X)
write.csv(air, "air.csv", row.names = FALSE)

questions <- character(6)
questions[1] <- round(SD<-sd(X),2)
questions[2] <- round(1.1*SD,2)
questions[3] <- round(0.9*SD,2)
questions[4] <- round(1.2*SD,2)
questions[5] <- round(1.15*SD,2)
questions[6] <- round(0.85*SD,2)
Drop=1+sample(5,1)
questions=questions[-Drop]

solutions <- c(TRUE,FALSE,FALSE,FALSE,FALSE)
explanations <- character(5)
explanations[1] <- "Answer found using sd()."
explanations[2:5] <- "Investigate use of sd() for finding standard deviations."

## permute order of solutions/questions
Order <- sample(1:5)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

Download the data file [air.csv](air.csv), the estimated amount of carbon monoxide (`CO`) in `r N` air samples by right clicking and 'Save link as', then load it into RStudio using `read_csv()` or the Import Dataset button.

Use `dplyr` to find the standard deviation of the estimated carbon monoxide. Which answer is closest?

```{r questionlist, echo = FALSE, results = "asis"}
library(tidyverse)
answerlist(questions, markup = "markdown")
```

Solution
========

```{r air_iqr, echo = TRUE, message=FALSE, comment=""}
air = read_csv("air.csv")
air |> summarise(sd(CO))
```

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================

extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice

