```{r data generation, echo = FALSE, results = "hide"}
library(dplyr)
library(readr)
library(exams)

odis <- read_csv("https://www.massey.ac.nz/~jcmarsha/161122/data/ODIs.csv")

# choose the variable(s)
var1 <- "Team"
var2 <- "Year"

answer <- 0
while (answer == 0) {
  # OK, now for each one, find the one we want to filter to
  options1 <- odis |> pull(var1) |> unique()
  options2 <- odis |> pull(var2) |> unique()

  selected1 <- sample(options1, 1)
  selected2 <- sample(options2, 1)

  answer_df <- odis |> filter(.data[[var1]] == selected1,
                  .data[[var2]] == selected2) |>
    count(Result)

  if (nrow(answer_df) > 1) {
    result <- sample(answer_df$Result, 1)

    answer <- answer_df |> mutate(Prop = n/sum(n)) |>
      filter(Result == result) |>
      pull(Prop)
  }
}

questions <- character(5)
questions[1] <- round(answer, 3)
questions[2] <- round(1-answer, 3)
questions[3] <- round(answer/2, 3)
questions[4] <- round(min(answer*2, 1),3)
questions[5] <- round((1-answer)/2, 3)
questions <- unique(questions)

solutions <- questions == round(answer, 3)
explanations <- character(length(solutions))

## permute order of solutions/questions
Order <- sample(seq_along(questions))
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

A dataset of one day cricket match results from 1971 to 2019 may be downloaded into RStudio using the following code:

```{r, eval=FALSE}
library(tidyverse)
odis <- read_csv("https://www.massey.ac.nz/~jcmarsha/161122/data/ODIs.csv")
```

Using these data, the **proportion** of games `r result` by `r selected1` in the year `r selected2` is:

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

```{r, echo=FALSE, message=FALSE, warning=FALSE, comment = "", results='asis'}
src <- knitr::knit_expand(text=c('```{r, message=FALSE, warning=FALSE}',
                   'library(tidyverse)',
                   'odis <- read_csv("https://www.massey.ac.nz/~jcmarsha/161122/data/ODIs.csv")',
                   'odis |>',
                   '  filter({{var1}} == "{{selected1}}",',
                   '         {{var2}} == "{{selected2}}") |>',
                   '  count(Result) |>',
                   '  mutate(Proportion = n/sum(n))',
                   '```'))
res = knitr::knit_child(text = src, options = list(comment = ""), quiet=TRUE)
cat(res)
```

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================

extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice
