```{r data generation, echo = FALSE, results = "hide"}
library(dplyr)
library(readr)
library(exams)

N=13+sample(10,1)
Min=20+sample(5,1)
Scalar=5+sample(4,1)
X=round(Min + Scalar*runif(N), 2)

air <- data.frame(CO=X)
write.csv(air, "air.csv", row.names = FALSE)

TheIQR=IQR(air$CO)

questions <- character(5)
questions[1] <- round(TheIQR,2)
questions[2] <- round(1.1*TheIQR,2)
questions[3] <- round(0.9*TheIQR,2)
questions[4] <- round(1.2*TheIQR,2)
questions[5] <- round(1.15*TheIQR,2)

solutions <- c(TRUE,FALSE,FALSE,FALSE,FALSE)
explanations <- character(5)
explanations[1] <- ""
explanations[2] <- ""
explanations[3] <- ""
explanations[4] <- ""
explanations[5] <- ""

## permute order of solutions/questions
Order <- sample(1:5)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

Download the data file [air.csv](air.csv) of the estimated amount of carbon monoxide (`CO`) in `r N` air samples by right clicking and 'Save link as', then load it into RStudio using `read_csv()` or the Import Dataset button.

Use `dplyr` to compute the interquartile range of carbon monoxide amounts. Which answer is closest?

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

```{r air_iqr, echo = TRUE, message=FALSE, comment=""}
library(tidyverse)
air = read_csv("air.csv")
air |> summarise(iqr = IQR(CO))
```

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================

extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice

