```{r data generation, echo = FALSE, results = "hide"}
library(dplyr)
library(readr)
library(exams)

schools = read_csv("https://www.massey.ac.nz/~jcmarsha/data/schools/schools.csv")

# choose the variable(s)

answer <- 0
while (answer == 0) {

  # Find the school we want
  options1 <- schools |> pull(RegionalCouncil) |> unique()
  regional_council <- sample(options1, 1)

  answer_df <- schools |> filter(RegionalCouncil == regional_council) |>
    count(SchoolGender)

  if (nrow(answer_df) > 1) {
    school_gender <- sample(answer_df$SchoolGender, 1)

    answer <- answer_df |> mutate(Prop = n/sum(n)) |>
      filter(SchoolGender == school_gender) |>
      pull(Prop)
  }

}

questions <- character(5)
questions[1] <- round(answer, 3)
questions[2] <- round(1-answer, 3)
questions[3] <- round(answer/2, 3)
questions[4] <- round(min(answer*2, 1),3)
questions[5] <- round((1-answer)/2, 3)
questions <- unique(questions)

solutions <- questions == round(answer, 3)
explanations <- character(length(solutions))

## permute order of solutions/questions
Order <- sample(seq_along(questions))
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]
```

Question
========

Using the `schools` data that was used in workshop A07, find the **proportion** of `r school_gender` schools in the `r regional_council` regional council.

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

```{r, echo=FALSE, message=FALSE, comment = "", results='asis'}
src <- knitr::knit_expand(text=c('```{r, message=FALSE}',
                   'schools <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/schools/schools.csv")',
                   'schools |>',
                   '  filter(RegionalCouncil == "{{regional_council}}") |>',
                   '  count(SchoolGender) |>',
                   '  mutate(Proportion = n/sum(n))',
                   '```'))
res = knitr::knit_child(text = src, options = list(comment = ""), quiet=TRUE)
cat(res)
```

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================

extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice

