```{r, echo=FALSE, results="hide"}
## DATA GENERATION
x=(runif(500)+runif(500)+runif(500))*20
MeanX=round(mean(x),1)
SDX=round(sd(x),1)
## QUESTION/ANSWER GENERATION
questions <- numeric(5)
questions[1] <- SDX
questions[2] <- MeanX
questions[3] <- MeanX+SDX
questions[4] <- 3*SDX
questions[5] <- round(SDX/4,2)

solutions <- c(TRUE,FALSE,FALSE,FALSE,FALSE)

questions <- masseystat::make_unique(questions, solutions, tolerance=0.05)

explanations <- character(5)

## permute order of solutions/questions
Order <- sample(1:5)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

The following histogram shows a sample of 500 results.

```{r, fig.width=7, fig.height=5, fig.cap="", fig.retina=2, echo=FALSE}
library(tidyverse)
ggplot(data.frame(x=x)) + geom_histogram(aes(x), bins=15)
```

The standard deviation of this data is likely to be close to:

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

Recall that for a Normal distribution, around 95% of the data is within 2 standard deviations of the mean, and around 99.5% of the data is within 3 standard deviations of the mean.

So, for a normal distribution we'd expect the range of the data to be around 4-6 standard deviations.

This also applies (approximately at least!) to distributions that are relatively normal, like this one. So take the range and divide by 4 to 6 to get an idea as to what the standard deviation might be.

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================
extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple choice

