---
title: "Workshop A01: Our first R script"
output: html_document
---

Our goal in this first set of exercises is to produce some plots and learn a bit about how R/RStudio works.

## Before we start

Make sure you have both R and RStudio installed. In addition, make sure you have the `tidyverse` package installed.

Both of these are detailed here:

https://r-resources.massey.ac.nz/help/usingrin161.122.html

## Loading up our first R script

Start by downloading labA01.R to somewhere on your computer. The easiest way to do this is to right click on the URL below and "Save file as..."

https://www.massey.ac.nz/~jcmarsha/161122/labs/labA01.R

Then load it into RStudio. The fail-safe way to do that is to first load RStudio up, and then use the File->Open File... option (alternatively the folder button on the toolbar), browsing to where the file was saved, and loading it up.

Alternatively, you may be able to right-click on the file in explorer or Finder, then Open with... and choose RStudio.

When done, you'll see that RStudio loads the script (a text file with a list of comments and commands) into the `Editor` pane in the top right. There's 3 other panes available: The `Console` pane in bottom left, which is where commands are run and (text) output given, The `Environment/History` pane in top right, and the `Files/Plots/Packages/Help` pane in bottom right. We'll use all of them today!

## Making sure it all works

1. Start by reading through the script. You'll see it has a bunch of comments at the top (each line beginning with a hash character) which explain what is going on, followed by some lines of code that perform various tasks.

2. The first command is `library(tidyverse)`. This tells RStudio to load in the libraries (also known as packages) called the `tidyverse`. Packages are a way to extend the base R/RStudio functionality. In this case, the tidyverse package is a set of packages for doing data analysis in a consistent way. See http://www.tidyverse.org for more information.

3. Place your cursor on the line `library(tidyverse)` and then click on the `Run` button in the top right of the editor pane. Alternatively, `Ctrl`-`Enter` (`Command`-`Enter` on a Mac) will run the line. What should happen is the line will be copied down to the `Console` window, where it will be executed. In this case, the package will be loaded and you should get a message that looks something like this:

```{r, echo=FALSE}
library(tidyverse)
```

If you get an error here, it's probably due to the `tidyverse` packages not being installed - check https://r-resources.massey.ac.nz/help/usingrin161.122.html or contact your lecturers to get things figured out.

## Now for something useful!

4. Now, we're going to load up some data on earthquakes from Fiji. This dataset is included with R/RStudio. Place your cursor on the `data(quakes)` line and run it. Again, the code will be copied down to the Console and executed. You'll see you now have a `quakes` object in the `Environment` pane in the top left.

5. Double click on the `quakes` object in the `Environment` pane to have a look - this shows you the data in a spreadsheet form. You should see 5 columns of data with observations from 1000 earthquakes.

6. The next 2 lines of code produce a plot. This seems quite a lot of effort! But that effort will be rewarded by allowing us to produce very elegant charts that tell clear stories later on!

    To run the two lines, you can place your cursor anywhere within those 2 lines and hit the `Run` button or `Ctrl`-`Enter` (`Command`-`Enter` on Mac). You can also highlight the 2 lines and do the same thing if you prefer. Notice that R/RStudio is treating the 2 lines as if they were a single statement - in fact, they are! The two parts are joined up using the `+` operator in this case. Essentially we "add" the `geom_point` bit on to the `ggplot` bit to produce the plot.

7. Let's break down what is going on, as it's quite a lot! One way to do that is to run it in sections.

    Type in (or copy + paste) `ggplot(data=quakes)` on a separate line in the R script and run that line.

    This tells R/RStudio to create a new plot using `quakes` as the data - it'll just give a blank canvas in the plot window.

    **NOTE** If you accidentally added the `+` from the original, when you run it you'll find that it copies to the `Console` but you'll be left with a `+ ` prompt there, rather than the usual `> ` prompt. This is R/RStudio knowing that the command is not complete, so it's waiting for the rest of it. If this happens and you don't intend it, just hit `Esc` key until the prompt returns to the usual `> `.

    Next, type in the bit inside the `geom_point` function: `aes(x=long, y=lat)` on a separate line in the R script and run that. You should get:

    ```{r, echo=FALSE}
    aes(x=long, y=lat)
    ```

    defining the mapping from "Aesthetic" features of the plot (e.g. the x and y axes) to features of the data. The `aes` is short for aesthetics.

    Lastly, the `geom_point()` command adds a layer of geometry to the plot, mapping from the data to the axes using the `aes` bit. In this case we're adding points. So this will add each observation as a point on the plot.

## Adding to our plot

In addition to the `x` and `y` aesthetics we've used above, there's a bunch of others. For example, you can change `colour`, `shape`, `size` as well as transparency of points using `alpha`. Plus lots more.

8. Try the following:

    ```{r, eval=FALSE}
    ggplot(data=quakes) +
      geom_point(mapping=aes(x=long, y=lat), col='pink', size=2)
    ```

    Notice the `'pink'` there is in quotation marks. This is because it's a label for something, not a quantity that R/RStudio would already know about. You can use either single quotes `'` or double quotes `"` as long as you're consistent (i.e. start and end with the same thing).
    
    Also, notice that when we specify multiple parameters to the function `geom_point()` we separate them using a comma. This is a general principle: functions in R/RStudio always use parentheses when we run them, and parameters that we specify are always separated by commas. Note we did that with the `aes()` function too.
    
9. You might want to experiment with the other parameters. What happens when you use `shape=2` for example? Or `alpha=0.4`? Try experimenting!

10. In the same way that `x` and `y` are mapped to features in our data, we can do the same for the other aesthetics. Try adding `col=mag` to the `aes()` function so that it looks like:

    ```{r, eval=FALSE}
    ggplot(data=quakes) +
      geom_point(aes(x=long, y=lat, col=mag))
    ```

11. You might want to try looking at `col=stations` as well. You should find they look really similar. This makes sense - larger earthquakes are likely felt more widely, so will be detected at more stations.

12. Once you're done, remember you can add comments to your script - save it in a safe place so you can go back to it later.

