```{r data_generation, echo=FALSE, results='hide'}

library(exams)

num_choice <- 4
questions_combn <- lapply(0:num_choice, function(x) combn(num_choice,x))

questions <- character(num_choice^2)
k<-1
for (i in 1:length(questions_combn)){
  candidate <- questions_combn[[i]]
  for(j in 1:ncol(candidate)){
      choice<-unique(candidate[,j])
      questions[k]<-paste0(choice, collapse = ', ')
      k<-k+1
  }
}

questions[1]<-paste('None of',paste0(1:num_choice, collapse = ', '))

solutions <- rep(FALSE,num_choice^2)

solutions[10] <- TRUE


explanations <- rep("",num_choice^2)


## permute order of solutions/questions
Order <- sample(c(which(solutions==TRUE),sample(which(solutions==FALSE),4)))
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

The adjusted $R^2$ accounts for the balance between which of the following items

1. the number of dependent variables in the model
2. the number of independent variables in the model
3. unusually large predictors
4. the goodness of fit of the model

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

The adjusted $R^2$ balances model fit ($R^2$) against model complexity (number of independent variables).

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================
  
extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice
