```{r data_generation, echo=FALSE, results='hide'}
library(exams)

questions <- character(4)
questions[1] <- "Residual versus fits plot"
questions[2] <- "Scale-location plot"
questions[3] <- "Scatter plot"
questions[4] <- "Q-Q plot"

solutions <- c(FALSE,FALSE,FALSE,TRUE)
explanations <- character(5)
explanations[1] <- ""
explanations[2] <- ""
explanations[3] <- ""
explanations[4] <- "Q-Q plot can only check the normality."

## permute order of solutions/questions
Order <- sample(1:4)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

Which of the following diagnostic tools can **not** be used to check the linearity and/or equal variance assumptions of a fitted linear model?

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

A scatterplot can potentially be used (is there a curve) but often a residual versus fit plot is more useful. The scale location plot can be used to assess equal variation as well. The normal Q-Q plot only assesses normality.
  
```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================
  
extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice
