```{r data generation, echo = FALSE, results = "hide"}
library(tidyverse)
library(exams)

Linearity <- sample(c(TRUE,FALSE),1)
EqualVar <- sample(c(TRUE,FALSE),1)
n <- 100
d <- data.frame(x = sort(runif(n, -2, 2)))
a <- sample(c(0, 1), 1)
b <- sample(c(-1, 1), 1)
Var.Factor<-1
if(EqualVar==FALSE) Var.Factor <- (1:n)/20
e <- rnorm(n, sd = 0.25*Var.Factor)
d$y <- a + b * d$x + e 
if(Linearity==FALSE) d$y <- a + b *exp( d$x) + e 

write.csv(d, "diagnostics.csv", row.names = FALSE, quote = FALSE)

questions <- character(6)
questions[1] <- "Linearity"
questions[2] <- "Normality"
questions[3] <- "Independence"
questions[4] <- "Equal variance"
questions[5] <- "Linearity and equal variance"
questions[6] <- "None of the above"

solutions <- c(!Linearity & EqualVar,FALSE,FALSE,Linearity & !EqualVar,!Linearity & !EqualVar,
               Linearity & EqualVar)

explanations <- character(6)
explanations[1] <- ""
explanations[2] <- ""
explanations[3] <- ""
explanations[4] <- ""
explanations[5] <- ""
explanations[6] <- ""

## permute order of solutions/questions
Order <- c(sample(1:5),6)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```    

Question
========

Download the data file [diagnostics.csv](diagnostics.csv) by right clicking and 'Save link as', then load it into RStudio using `read_csv()` or the Import Dataset button.

Use `lm()` to fit a linear model to `x` and `y` and perform standard residual diagnostics on your model. Which one of the following assumptions fails to hold based on your diagnostics?
  
```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

To replicate the analysis in R:

```{r, eval=FALSE}
library(tidyverse)
d <- read_csv("diagnostics.csv")
d %>% ggplot(aes(x=x,y=y)) + geom_point()
m <- lm(y ~ x, data = d)
plot(m)
```

Your residuals versus fitted plot should look like this:

```{r, echo=FALSE, fig.cap=""}
plot(lm(y ~ x, data = d), which=1, add.smooth = FALSE)
```

So your conclusion should be that linearity is `r ifelse(Linearity, "", "not ")`satisfied and equal variance is `r ifelse(EqualVar, "", "not ")`satisfied.

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================
extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice