```{r data generation, echo = FALSE, results = "hide"}
library(tidyverse)
library(exams)


n <- 400


Brand_effect <- FALSE
 
Inter_effect <- TRUE

Humidity<- runif(n,0,1)

Brand <- c("Gluten",sample(c("Gluten","Guru","Gluey"),n-1,replace = TRUE))

mean <- 1.5 - Humidity + Brand_effect*(0.5*(Brand=='Guru') + 1*(Brand=='Gluey')) + 
  Inter_effect*(0.5 *(Brand=='Guru') - 0.5 *(Brand=='Gluey'))*Humidity 

Lifetime <- exp(rnorm(n,mean,sd=0.01))

d <- data.frame(Lifetime,Humidity,Brand)

write.csv(d, "glue.csv", row.names = FALSE, quote = FALSE)

questions <- character(5)
questions[1] <- "A log transformation on lifetimes is needed when modelling."

questions[2] <- "The humidity condition is always significant."

questions[3] <- "The interaction effects are significant after accounting for the brand effects and the humidity."

questions[4] <- "The brand effects are significant after accounting for the humidity condition."

questions[5] <- "The brand effects are significant after accounting for both the humidity conditions and the interaction effects."

solutions <- c(FALSE,FALSE,FALSE,FALSE,TRUE)

explanations <- rep("",length(solutions))

## permute order of solutions/questions
Order <- sample(c(sample(which(solutions!=TRUE),4),which(solutions==TRUE)))
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```    

Question
========

Download the data file [glue.csv](glue.csv) by right clicking and 'Save link as', then load it into RStudio using `read_csv()` or the Import Dataset button. The data set summarises the lifetimes of adhesive bonds (`Lifetime`) produced by three manufacturers (`Brand`:`Gluey`,`Guru`,`Gluten`) under different humidity conditions (`Humidity`). 

Fit suitable linear models to this data set, peruse their ANOVA tables and R summaries, and perform residual diagnostics. Based on your analysis, which one of the following statements is **not** true at the significance level 0.05?

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

To replicate the analysis in R:

```{r, eval=FALSE}
library(tidyverse)
d <- read_csv('glue.csv')

# Try a full model. It should be obvious that a log transform is needed
full.mod <- lm(Lifetime ~ Brand + Humidity + Brand:Humidity,data=d)
plot(full.mod)

# Fit some models with the log transform
simple.mod.log <- lm(log(Lifetime) ~ Humidity, data=d)
partial.mod.log <- lm(log(Lifetime) ~ Brand + Humidity, data=d)
partial.mod.inter.log <- lm(log(Lifetime) ~ Humidity + Brand:Humidity, data=d)
full.mod.log <- lm(log(Lifetime) ~ Brand + Humidity + Brand:Humidity, data=d)

# Assess these models
anova(full.mod.log)
anova(simple.mod.log,partial.mod.log)
anova(partial.mod.log,full.mod.log)
anova(partial.mod.inter.log,full.mod.log)
summary(partial.mod.inter.log)
summary(full.mod.log)
```

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================
extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice