```{r data generation, echo = FALSE, results = "hide"}
library(tidyverse)
library(exams)


n <- 400

Brand_effect <- FALSE
 
Inter_effect <- TRUE

Humidity<- runif(n,0,1)

Brand <- c("Gluten",sample(c("Gluten","Guru","Gluey"),n-1,replace = TRUE))

mean <- 1.5 - Humidity + Brand_effect*(0.5*(Brand=='Guru') + 1*(Brand=='Gluey')) + 
  Inter_effect*(0.5 *(Brand=='Guru') - 0.5 *(Brand=='Gluey'))*Humidity 

pvalue <- 0.05
while(pvalue<=0.05){
Lifetime <- exp(rnorm(n,mean,sd=0.01))
d <- data.frame(Lifetime,Humidity,Brand)
partial.mod.inter.log<-lm(log(Lifetime)~Brand:Humidity+Humidity,data=d)
full.mod.log<-lm(log(Lifetime)~Brand+Humidity+Brand:Humidity,data=d)
ANT<-anova(partial.mod.inter.log,full.mod.log)
pvalue<-ANT[2,6]
}

questions <- character(5)
questions[1] <- "A log transformation on lifetimes is needed in modelling."

questions[2] <- "The humidity condition is always significant."

questions[3] <- "The interaction effects are significant after accounting for the brand effects and the humidity."

questions[4] <- "The brand effects are significant after accounting for the humidity condition."

questions[5] <- "The brand effects are significant after accounting for both the humidity conditions and the interaction effects."


solutions <- c(FALSE,FALSE,FALSE,FALSE,TRUE)

explanations <- rep("",length(solutions))

## permute order of solutions/questions
Order <- sample(c(sample(which(solutions!=TRUE),4),which(solutions==TRUE)))
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```    

Question
========
A scientist would like to study the lifetimes of adhesive bonds (`Lifetime`) produced by three manufacturers (`Brand`:`Gluey`,`Guru`,`Gluten`) under different humidity conditions (`Humidity`). He performed the following regression analysis with R

```{r, fig.width=8, fig.height=5, fig.cap="", fig.retina=2}
full.mod<-lm(Lifetime~Brand+Humidity+Brand:Humidity,data=d)
par(mfrow=c(2,2), mar=c(4,4,2,2))
plot(full.mod)
simple.mod.log<-lm(log(Lifetime)~Humidity,data=d)
partial.mod.log<-lm(log(Lifetime)~Brand+Humidity,data=d)
partial.mod.inter.log<-lm(log(Lifetime)~Brand:Humidity+Humidity,data=d)
full.mod.log<-lm(log(Lifetime)~Brand+Humidity+Brand:Humidity,data=d)
anova(full.mod.log)
anova(simple.mod.log,partial.mod.log)
anova(partial.mod.log,full.mod.log)
anova(partial.mod.inter.log,full.mod.log)
summary(partial.mod.inter.log)
summary(full.mod.log)
plot(partial.mod.inter.log)
```

Peruse the above ANOVA tables, R summaries, and residual diagnostics plots. Based on your observation, which one of the following statements is **not** true at the significance level 0.05?

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================
extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice