```{r data_generation, echo=FALSE, results='hide'}
library(exams)

questions <- character(5)
questions[1] <- "1"
questions[2] <- "2"
questions[3] <- "3"
questions[4] <- "None of 1,2,3,4"

solutions <- c(FALSE,FALSE,TRUE,FALSE)
explanations <- character(4)

## permute order of solutions/questions
Order <- sample(1:4)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

If the *F* test statistic for a regression is greater than the critical value from the *F* distribution, it implies that

1. none of the independent variables in the regression model have a significant effect on the dependent variable.
 
2. all of the independent variables in the regression model have significant effects on the dependent variable.

3. one or more of the independent variables in the regression model have a significant effect on the dependent variable.

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

The *F* statistic is the ratio of the variance explained by the model to the variance in the residuals:

$$
F = \frac{\mbox{variance explained}}{\mbox{variance unexplained}}
$$

So if $F$ is larger than what we'd expect by chance (the critical value) our conclusion is that some variance in the outcome is explained, so at least some of our independent variables must be useful for explaining the dependent variable.

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================
  
extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice
