```{r data generation, echo = FALSE, results = "hide"}
library(tidyverse)
library(exams)

which_coef <- sample(c("intercept", "slope"), 1)

d <- data.frame(x = runif(100, -2, 2))
a <- sample(c(0, 1), 1)
b <- sample(c(-1, 0, 1), 1) 
d$y <- a + b * d$x + rnorm(100, sd = 0.25)

write.csv(d, "regression2.csv", row.names = FALSE, quote = FALSE)

m <- lm(y ~ x, data = d)
ahat <- coef(m)[1]
bhat <- coef(m)[2]
ahatSE <- summary(m)$coefficients[1, 2]
bhatSE <- summary(m)$coefficients[2, 2]

intervals<-round(rbind(cbind(ahat-(1:3)*ahatSE,ahat+(1:3)*ahatSE),
      cbind(bhat-(1:3)*bhatSE,bhat+(1:3)*bhatSE)),3)

questions <- paste0('(',intervals[,1],',',intervals[,2],')')
  
solutions <- c(FALSE,which_coef == "intercept",FALSE,FALSE,which_coef == "slope",FALSE)

explanations <- character(6)
explanations[1] <- ""
explanations[2] <- ""
explanations[3] <- ""
explanations[4] <- ""
explanations[5] <- ""
explanations[6] <- ""

## permute order of solutions/questions
Candidate <- 1:6
TrueCandidate <- Candidate[solutions==TRUE]
Order <- c(TrueCandidate, sample(Candidate[-TrueCandidate],4))
possible_answers <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```    

Question
========

Download the data file [regression2.csv](regression2.csv) by right clicking and 'Save link as', then load it into RStudio using `read_csv()` or the Import Dataset button.

Use `lm()` to fit a linear model to `x` and `y` and make a summary of the model. Which one of the following intervals is the 95% confidence interval for the estimated `r which_coef`?

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(possible_answers, markup = "markdown")
```

Solution
========

To replicate the analysis in R:

```
library(tidyverse)
library(broom)
d <- read_csv("regression2.csv")
m <- lm(y ~ x, data = d)
tidy(m) |>
  mutate(low = estimate - 2*std.error,
         high = estimate + 2*std.error)
```

You should get:

```{r, echo=FALSE}
broom::tidy(m) |>
  mutate(low = estimate - 2*std.error,
         high = estimate + 2*std.error)
```


```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================
extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice