```{r data generation, echo = FALSE, results = "hide"}
library(tidyverse)
library(exams)
d <- data.frame(x = runif(100, -1, 1))
a <- 0
b <- sample(c(-1, 1), 1) * sample(c(0, 0.6, 0.9), 1)
d$y <- a + b * d$x + rnorm(100, sd = 0.25)

write.csv(d, "regression3.csv", row.names = FALSE, quote = FALSE)

m <- lm(y ~ x, data = d)
# bhat <- coef(m)[2]
# bpvl <- summary(m)$coefficients[2, 4]
# bsol <- c(bpvl >= 0.05, (bpvl < 0.05) & (bhat > 0), (bpvl < 0.05) & (bhat < 0))

newdata<-data.frame(x=c(-2,-1,0,1,2))

prediction <- round(predict(m,newdata),3)
confidenceinterval <- round(predict(m,newdata,interval ='confidence'),3)
predictioninterval <- round(predict(m,newdata,interval ='prediction'),3)
predictioninterval99 <- round(predict(m,newdata,interval ='confidence',level=0.99),3)

type <- sample(c("prediction", "95% confidence interval", "95% prediction interval", "99% confidence interval"), 1)

if (type == "prediction") {
  answer <- tibble(value = prediction,
                 thing = paste0(type, " at x=", newdata$x),
                 xval = newdata$x)
  answer_func <- "predict(m, newdata=newdata)"
} else if (type == "95% confidence interval") {
  answer <- tibble(value = paste0('(',confidenceinterval[,2],',',confidenceinterval[,3],')'),
                 thing = paste0(type, " at x=", newdata$x),
                 xval = newdata$x)
  answer_func <- "predict(m, newdata=newdata, interval='confidence')"
} else if (type == "95% prediction interval") {
  answer <- tibble(value = paste0('(',predictioninterval[,2],',',predictioninterval[,3],')'),
                 thing = paste0(type, " at x=", newdata$x),
                 xval = newdata$x)
  answer_func <- "predict(m, newdata=newdata, interval='prediction')"
} else {
  answer <- tibble(value = paste0('(',predictioninterval99[,2],',',predictioninterval99[,3],')'),
                 thing = paste0(type, " at x=", newdata$x),
                 xval = newdata$x)
  answer_func <- "predict(m, newdata=newdata, interval='confidence', level=0.99)"
}

#  min_diff <- answer |> pull(value) |> sort() |> diff() |> min()
#  if (min_diff / (answer |> pull(value) |> min()) > 0.2)
   
possible_answers <- answer |> pull(value)

possible_questions <- answer # |> filter(thing != "baseline")

which_question <- sample_n(possible_questions, 1)

solutions <- possible_answers == (which_question |> pull(value))

explanations <- paste0("That is the ", answer |> pull(thing), ".")

## permute order of solutions/questions
Order <- sample(1:5)
possible_answers <- possible_answers[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]
```    

Question
========
Download the data file [regression3.csv](regression3.csv) by right clicking and 'Save link as', then load it into RStudio using `read_csv()` or the Import Dataset button.

Use `lm()` to fit a linear model to `x` and `y` and make a prediction at x=`r which_question |> pull(xval)`.

Which of the following is a `r which_question |> pull(thing)`?

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(possible_answers, markup = "markdown")
```

Solution
========

To replicate the analysis in R:

```
d <- read.csv("regression3.csv")
m <- lm(y ~ x, data = d)
summary(m)
newdata <- data.frame(x = `r which_question |> pull(xval)`)
`r answer_func`
```

You should get

```{r, echo=FALSE}
newdata <- data.frame(x = which_question |> pull(xval))
eval(str2expression(answer_func))
```

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================
extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice