```{r data generation, echo = FALSE, results = "hide"}
library(tidyverse)
library(exams)

d <- data.frame(x = runif(100, -2, 2))
a <- sample(c(0, 1), 1)
b <- sample(c(-1, 0, 1), 1) 
d$y <- a + b * d$x + rnorm(100, sd = 0.25)

level <- sample(c(0.05,0.01), 1)

write.csv(d, "regression2.csv", row.names = FALSE, quote = FALSE)

m <- lm(y ~ x, data = d)
ahat <- coef(m)[1]
bhat <- coef(m)[2]
apvl <- summary(m)$coefficients[1, 4]
bpvl <- summary(m)$coefficients[2, 4]

questions <- character(6)
questions[1] <- "*x* and *y* are not significantly correlated but the baseline is significant."
questions[2] <- "None of the slope and baseline is significant."
questions[3] <- "*y* increases significantly with *x* with a significant baseline."
questions[4] <- "*y* decreases significantly with *x* with an insignificant baseline."
questions[5] <- "*y* increases significantly with *x* with an insignificant baseline."
questions[6] <- "*y* decreases significantly with *x* with a significant baseline."

solutions <- c( (bpvl >= level) & (apvl < level), 
                (bpvl >= level) & (apvl >= level),
                (bpvl < level) & (bhat > 0) & (apvl < level), 
                (bpvl < level) & (bhat < 0) & (apvl > level),
                (bpvl < level) & (bhat > 0) & (apvl > level), 
                (bpvl < level) & (bhat < 0) & (apvl < level)
                )

explanations <- character(6)
explanations[1] <- ""
explanations[2] <- ""
explanations[3] <- ""
explanations[4] <- ""
explanations[5] <- ""
explanations[6] <- ""

## permute order of solutions/questions
TrueCandidate <- which(solutions==TRUE)
Candidate <- 1:6
Order <- c(TrueCandidate, sample(Candidate[-TrueCandidate],4))
possible_answers <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  

```    

Question
========

Download the data file [regression2.csv](regression2.csv) by right clicking and 'Save link as', then load it into RStudio using `read_csv()` or the Import Dataset button.

Use `lm()` to fit a linear model to `x` and `y` and make a summary of the model. Which of the following statements is true about the fitted linear model at the significance level `r level`?

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(possible_answers, markup = "markdown")
```

Solution
========

To replicate the analysis in R:

```
d <- read.csv("regression2.csv")
m <- lm(y ~ x, data = d)
summary(m)
```

You should get:

```{r echo=FALSE}
summary(m)
```

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================
extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice