```{r data generation, echo = FALSE, results = "hide"}
library(tidyverse)
library(exams)

transformtype <- sample(c('linear', 'power', 'log', 'exp'), 1)

if (transformtype == 'power') {
  d <- data.frame(x = runif(100, 0.5, 2))
  a <- 1
  b <- sample(c(-1,1),1)
  d$y <- a* d$x^(3*b)*exp(rnorm(100, sd = 0.25))
} else if (transformtype == 'exp') {
  d <- data.frame(x = runif(100, 0.5, 2))
  a <- 1
  b <- sample(c(-1,1),1)
  d$y <- a*exp(2*b* d$x)*exp(rnorm(100, sd = 0.25))
} else if (transformtype == 'log') {
  d <- data.frame(x = runif(100, 0.1, 2))
  a <- 1
  b <- sample(c(-1,1),1)
  d$y <- a + b * log( d$x) + rnorm(100, sd = 0.1)
} else {
  d <- data.frame(x = runif(100, 0.5, 2))
  a <- 1
  b <- sample(c(-1,1),1)
  d$y <- a + b * d$x + rnorm(100, sd = 0.1)
}
write.csv(d, "regression.csv", row.names = FALSE, quote = FALSE)

m <- lm(y ~ x, data = d)

bhat <- coef(m)[2]

transform <- transformtype %in% c('power','exp','log')

questions <- character(6)
questions[1] <- "The scatter plot shows an increasing trend and the linear model fits it well."
questions[2] <- "The scatter plot shows a decreasing trend and the linear model fits it well."
questions[3] <- "The scatter plot shows an increasing trend but the linear model fits it poorly."
questions[4] <- "The scatter plot shows a decreasing trend but the linear model fits it poorly."
questions[5] <- "A transformation is needed here."
questions[6] <- "A transformation is not needed here."

solutions <- c( (bhat > 0)  & !transform, 
                (bhat < 0)  & !transform, 
                (bhat > 0)  & transform, 
                (bhat < 0)  & transform,
                transform, 
                !transform
                )

explanations <- character(6)
explanations[1] <- ""
explanations[2] <- ""
explanations[3] <- ""
explanations[4] <- ""
explanations[5] <- ""
explanations[6] <- ""

## permute order of solutions/questions
Candidate <- 1:6
TrueCandidate <- Candidate[solutions==TRUE]
Order <- c(sample(TrueCandidate,1), Candidate[-TrueCandidate])
possible_answers <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```    

Question
========

Download the data file [regression.csv](regression.csv) by right clicking and 'Save link as', then load it into RStudio using `read_csv()` or the Import Dataset button.

Use `lm()` to fit a linear model to `x` and `y` and visualise your fitted linear model.

Which of the following statements is true about the fitted linear model at the significance level 0.05?

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(possible_answers, markup = "markdown")
```

Solution
========

To replicate the analysis in R:

```
d <- read.csv("regression.csv")
m <- lm(y ~ x, data = d)
library(visreg)
visreg(m, gg=TRUE)
```

You should get:

```{r, echo=FALSE, fig.cap=""}
library(visreg)
visreg(m,gg=TRUE)
```

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================
extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice