```{r data generation, echo = FALSE, results = "hide"}
library(tidyverse)
library(exams)

n <- 100
p <- 4

ture_p <- sample(2:p,1)

x <- matrix(runif(n*p),ncol=p)

select <- sort(sample(1:p,ture_p))

X <- x[,select]

alpha <- 3

beta <- (2*rbinom(ture_p,1,0.5)-1)*2

y <- X%*%beta + alpha + rnorm(n)

data<-data.frame(y,x)

write.csv(data, "regression.csv", row.names = FALSE, quote = FALSE)

questions <- character()
adj.r2 <- numeric()

for(eff_p in 1:p){
x_combn <-  combn(p,eff_p)
adj.r2 <-  c(adj.r2,apply(x_combn,MARGIN=2,function(t) summary(lm(y~x[,t]))$adj.r.squared))
x_names <- apply(x_combn,MARGIN=2,
                function(t) colnames(data.frame(x))[t])
if(eff_p==1) x_names <- t(x_names)
questions <- c(questions,apply(x_names,MARGIN=2,
                function(t) paste0('y~',paste0(t,collapse='+'))))
}

solutions <- adj.r2==max(adj.r2)

explanations <- rep("",length(solutions))

## permute order of solutions/questions
Order <- sample(c(sample(which(solutions!=TRUE),4),which(solutions==TRUE)))
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```    

Question
========

Download the data file [regression.csv](regression.csv) by right clicking and 'Save link as', then load it into RStudio using `read_csv()` or the Import Dataset button. 

Which one of the following regression formulae yields the best compromise between the complexity and the goodness of fit among the corresponding fitted linear models, assessed via adjusted $R^2$?

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

The following R code **can be modified** to replicate the full analysis in R. You can then compare the 

```{r, eval=FALSE}
library(tidyverse)
library(broom)
d <- read_csv('regression.csv')

# fit models - modify this for the models you want to fit.
mod1 <- lm(y~X1, data=d)
mod2 <- lm(y~X1+X2, data=d)
mod3 <- lm(y~X1+X2+X3, data=d)
mod4 <- lm(y~X1+X2+X3+X4, data=d)

# check the adj.r.squared. A nice way to do this is with broom::glance
bind_rows(glance(mod1),
          glance(mod2),
          glance(mod3),
          glance(mod4), .id="model")
```

You'll get something like:

```{r, echo=FALSE}
map(questions, ~ broom::glance(lm(as.formula(.), data=data))) %>%
  set_names(questions) %>%
  bind_rows(.id='model')
```

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================
extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice