```{r data generation, echo = FALSE, message=FALSE, results = "hide"}
library(tidyverse)
library(exams)

n <- 100
p <- 3

ture_p <- sample(2:p,1)

x <- matrix(runif(n*p),ncol=p)

select <- sort(sample(1:p,ture_p))

X <- x[,select]

alpha <- 3

beta <- (2*rbinom(ture_p,1,0.5)-1)*2

y <- X%*%beta + alpha + rnorm(n)

d <- data.frame(y,x)

questions <- character()
adj.r2 <- numeric()

for(eff_p in 1:p){
  x_combn <-  combn(p,eff_p)
  adj.r2 <-  c(adj.r2,apply(x_combn,MARGIN=2,function(t) summary(lm(y~x[,t]))$adj.r.squared))
  x_names <- apply(x_combn,MARGIN=2,
                  function(t) colnames(data.frame(x))[t])
  if(eff_p==1) x_names <- t(x_names)
  questions <- c(questions,apply(x_names,MARGIN=2,
                  function(t) paste0('y ~ ',paste0(t,collapse=' + '))))
}

solutions <- adj.r2==max(adj.r2)

explanations <- rep("",length(solutions))

## permute order of solutions/questions
Order <- sample(c(sample(which(solutions!=TRUE),4),which(solutions==TRUE)))
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

Several linear models were fitted to some data `y`, `X1`, `X2`, and `X3`, with the results below.

```{r generating, echo=FALSE, message=FALSE, comment = "", results='asis'}
src <- knitr::knit_expand(text=c('```{r summaryoutput, message=FALSE}',
                   'summary(lm({{questions[1]}}, data=d))',
                   'summary(lm({{questions[2]}}, data=d))',
                   'summary(lm({{questions[3]}}, data=d))',
                   'summary(lm({{questions[4]}}, data=d))',
                   'summary(lm({{questions[5]}}, data=d))',
                   '```'))
res = knitr::knit_child(text = src, quiet=TRUE)
cat(res)
```

Which one of the following regression formulae yields the best compromise between the complexity and the goodness of fit among the corresponding fitted linear models?

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

The adjusted $R^2$ should be used.

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================
extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice