```{r data generation, echo = FALSE, results = "hide"}
library(tidyverse)
library(exams)
library(broom)
NO <- sample(c(0,5),1)
cutoff<-sample(c(0.1,0.25), 1)
n <-50
d <- data.frame(x = sort(runif(n, -2, 2)))
a <- sample(c(0, 1), 1)
b <- sample(c(-1, 1), 1)
e <- rnorm(n, sd = 0.25) 
d$y <- a + b * d$x + e 

OL <- sample(1:n,NO)

d$y[OL] <- sample(c(-(3:6),3:6),NO)
d$x[OL] <- sample(c((-4):4),NO)

m <- lm(y ~ x, data = d)

OL.detected <- augment(m) |> filter(.cooksd>cutoff) |> summarise(n()) |> as.numeric()
possible.questions <- OL.detected+c(-2,-1,0,1,2)
if (min(possible.questions)<0) possible.questions <- possible.questions - min(possible.questions)

write.csv(d, "regression2.csv", row.names = FALSE, quote = FALSE)
questions <- as.character(possible.questions)
solutions <- possible.questions==OL.detected

explanations <- character(5)
explanations[1] <- ""
explanations[2] <- ""
explanations[3] <- ""
explanations[4] <- ""
explanations[5] <- ""

## permute order of solutions/questions
Order <- sample(1:5)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]
```    

Question
========

Download the data file [regression2.csv](regression2.csv) by right clicking and 'Save link as', then load it into RStudio using `read_csv()` or the Import Dataset button.

Use `lm()` to fit a linear model to `x` and `y` and perform standard residual diagnostics on your model. Using a cut-off Cook's distance `r cutoff`, how many outliers you could find from this data set?
  
```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

To replicate the analysis in R:

```
# read and look at the data
library(tidyverse)
library(broom)
d <- read_csv("regression2.csv")
d |> ggplot(aes(x=x,y=y)) + geom_point()

# fit model and check for cooks distance past the cutoff.
m <- lm(y ~ x, data = d)
plot(m,which=4)

# find the outliers past the cutoff
augment(m) |> filter(.cooksd > `r cutoff`)
```

You should get:

```{r, echo=FALSE, fig.cap=""}
plot(m, which=4)
augment(m) |> filter(.cooksd > cutoff)
```

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================
extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice