```{r data_generation, echo=FALSE, results='hide'}
library(exams)

num_choice <- 7

questions <- character(num_choice)
questions[1] <- "numerical"
questions[2] <- "quantitative"
questions[3] <- "qualitative"
questions[4] <- "factor"
questions[5] <- "nominal"
questions[6] <- "categorical"
questions[7] <- "All of the above"

solutions <- c(FALSE,FALSE,FALSE,FALSE,FALSE,FALSE,TRUE)
explanations <- character(num_choice)
explanations[1] <- ""
explanations[2] <- ""
explanations[3] <- ""
explanations[4] <- ""
explanations[5] <- ""
explanations[6] <- ""
explanations[7] <- ""

## permute order of solutions/questions
Order <- c(sample(1:6,4),7)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

The predictor variables *x*'s in a linear model can be 

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

We can always code categorical (non-numeric) variables as numeric by introducing binary variables for the different categories, so both categorical and numeric variables can be included as predictors.

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================
  
extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice
