```{r data generation, echo = FALSE, results = "hide"}
library(tidyverse)
library(exams)

Normality <- FALSE

n <- 200
d <- data.frame(x = sort(runif(n, -2, 2)))
a <- sample(c(0, 1), 1)
b <- sample(c(-1, 1), 1)
e <- rnorm(n, sd = 0.25) 
if (Normality != TRUE) e <- rexp(n,1) -1 
d$y <- a + b * d$x + e 


write.csv(d, "regression2.csv", row.names = FALSE, quote = FALSE)

questions <- character(5)
questions[1] <- "Linearity"
questions[2] <- "Normality"
questions[3] <- "Independence"
questions[4] <- "Equal variance"
questions[5] <- "None of the above"

solutions <- c(FALSE,TRUE,FALSE,FALSE,FALSE)

explanations <- character(5)
explanations[1] <- ""
explanations[2] <- ""
explanations[3] <- ""
explanations[4] <- ""
explanations[5] <- ""

## permute order of solutions/questions
Order <- c(sample(1:4),5)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```    

Question
========

Download the data file [regression2.csv](regression2.csv) by right clicking and 'Save link as', then load it into RStudio using `read_csv()` or the Import Dataset button.

Use `lm()` to fit a linear model to `x` and `y` and perform standard residual diagnostics on your model. Which one of the following assumptions fails to hold based on your diagnostics?
  
```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========
  
To replicate the analysis in R:

```{r, eval=FALSE}
# read in and look at the data
library(tidyverse)
d <- read_csv("regression2.csv")
d |> ggplot(aes(x=x,y=y)) + geom_point()

# fit model
m <- lm(y ~ x, data = d)
plot(m)
```

You should find that your diagnostic plots look like this:

```{r, echo=FALSE, fig.cap=""}
par(mfrow=c(2,2), mar = c(4,4,2,2))
plot(lm(y~x, data=d), add.smooth=FALSE)
```

and hence normality is clearly not satisfied.

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================
extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice