```{r data_generation, echo=FALSE, results='hide'}
library(exams)

questions <- character(5)
questions[1] <- "Cook's distance"
questions[2] <- "Residuals vs leverage plot"
questions[3] <- "R squared"
questions[4] <- "Box plot"
questions[5] <- "Scatter plot"

solutions <- c(FALSE,FALSE,TRUE,FALSE,FALSE)
explanations <- character(5)
explanations[1] <- ""
explanations[2] <- ""
explanations[3] <- "R squared only gives a summary on the goodness of fit."
explanations[4] <- ""
explanations[5] <- ""

## permute order of solutions/questions
Order <- c(sample(1:5))
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========

Which of the following tools can **not** be used to identify potential outliers in a data set?
  
```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

Outliers might appear on boxplots or scatterplots. They can also be seen on a residuals vs leverage plot, or a cook's distance plot. They cannot be determined using $R^2$, which only tells you the proportion of variation explained by the model.

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================
  
extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice
