
```{r data_generation, echo=FALSE, results='hide'}
library(exams)

questions <- character(7)
questions[1] <- "Only 2"
questions[2] <- "Only 1"
questions[3] <- "Only 3"
questions[4] <- "1 and 2"
questions[5] <- "1 and 3"
questions[6] <- "2 and 3"
questions[7] <- "1,2 and 3"

solutions <- c(TRUE,FALSE,FALSE,FALSE,FALSE,FALSE,FALSE)
explanations <- character(7)
explanations[1] <- "A yearly time series usually exhibits no seasonal patterns. The frequency is just 1 within one year."
explanations[2] <- ""
explanations[3] <- ""
explanations[4] <- ""
explanations[5] <- ""
explanations[6] <- ""
explanations[7] <- ""

## permute order of solutions/questions
Order <- c(1,sample(2:7,4))
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========
  
Which of the following time series **may not** have any seasonal effects?

1. The number of patients hospitalised per month in NZ.
2. The number of babies born per year in NZ.
3. The daily precipitation in Palmerston North.

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

The number of patients hospitalised per month will likely be seasonal as some risks of hospitalisation (e.g. sport, recreation) will be seasonal. In addition, clearly daily precipitation will be seasonal (more rain in winter and spring). However, number of babies born per year is unlikely to be seasonal as we only have one
observation per year, and there is unlikely to be longer term seasonal behaviour (e.g. by decade).
  
```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================
  
extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice

