
```{r data_generation, echo=FALSE, results='hide'}
library(exams)

questions <- character(5)
questions[1] <- "Only 3"
questions[2] <- "1 and 2"
questions[3] <- "2 and 3"
questions[4] <- "1 and 3"
questions[5] <- "1,2 and 3"

solutions <- c(FALSE,FALSE,FALSE,FALSE,TRUE)
explanations <- character(5)
explanations[1] <- ""
explanations[2] <- ""
explanations[3] <- ""
explanations[4] <- ""
explanations[5] <- "All the examples have a regularly spaced time component."

## permute order of solutions/questions
Order <- sample(1:5)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========
  
Which of the following is an example of time series problem?

1. Estimating the number of rooms booking at Chateau Tongariro Hotel in next 6 months.
2. Estimating the total sales in the next 3 years of Harvey Norman.
3. Estimating the number of calls to ANZ customer service center for the next week.

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

Each of these examples has a time aspect to them (data collected over time), so all are time series problems.

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================
  
extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice

