
```{r data_generation, echo=FALSE, results='hide'}
library(exams)
 
questions <- character(5)
questions[1] <- "Fit a simple linear model to the data."
questions[2] <- "Calculate the mean and variance of your data."
questions[3] <- "Visualise your data on a graph."
questions[4] <- "Identify the important independent variables."
questions[5] <- "None of the above."

solutions <- c(FALSE,FALSE,TRUE,FALSE,FALSE)
explanations <- character(5)
explanations[1] <- ""
explanations[2] <- ""
explanations[3] <- "Visualisation is always the first step of your analysis."
explanations[4] <- ""
explanations[5] <- ""

## permute order of solutions/questions
Order <- c(sample(1:4),5)
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```

Question
========
  
After pre-processing your data set, the first step in time-series analysis (and any statistical analysis) is to

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

The first step is always exploratory data analysis, i.e. visualising the data. This helps to highlight
any issues that there may be (scales, transformations needed etc).

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================
  
extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice

