```{r data generation, echo = FALSE, results = "hide"}
library(tidyverse)
library(exams)

T<-96

setup_options <- tibble::tribble(~linear, ~seasonal, ~transform, ~error,
                                 TRUE, TRUE, FALSE, 1/4,
                                 FALSE, TRUE, FALSE, 1/20,
                                 FALSE, FALSE, FALSE, 1/10,
                                 TRUE, TRUE, TRUE, 1/2)

setup <- setup_options |> sample_n(1)

lineartrend <- setup |> pull(linear)
quadratictrend<-!lineartrend
seasonality <- setup |> pull(seasonal)
transformation <- setup |> pull(transform)
error_rate <- setup |> pull(error)

trigger <- c(lineartrend,quadratictrend,seasonality,transformation)

freq <- sample(c(4,12),1)
seasonal_max <- ifelse(lineartrend, 30, 200)
seasonalindex<- 0
if(seasonality==TRUE) seasonalindex <- rep(scale(purrr::rdunif(freq,seasonal_max),scale=FALSE),T/freq)

d <- data.frame(t = 1:T)
b <- 1
c <- 0.05
d$y <- b * d$t + seasonalindex + rexp(T,error_rate)
if(quadratictrend==TRUE)  d$y <- d$y + c * (d$t)^2 
if(transformation==TRUE)  d$y <- exp(d$y/30)

d <- d |> mutate(s=paste0("S",rep(1:freq,T/freq)))

write.csv(d, "timeseries.csv", row.names = FALSE, quote = FALSE)

possible <- c('Fit a linear trend to the time series.',
'Fit a quadratic trend to the time series.',
'Address the seasonalities via adding factors.',
'Take a log transformation on *y*.')

questions <- character(8)
questions[1] <- paste(possible[c(TRUE,FALSE,FALSE,FALSE)], collapse = " ")
questions[2] <- paste(possible[c(FALSE,TRUE,FALSE,FALSE)], collapse = " ")
questions[3] <- paste(possible[c(FALSE,FALSE,TRUE,FALSE)], collapse = " ")
questions[4] <- paste(possible[c(TRUE,FALSE,TRUE,FALSE)], collapse = " ")
questions[5] <- paste(possible[c(FALSE,TRUE,TRUE,FALSE)], collapse = " ")
questions[6] <- paste(possible[c(TRUE,FALSE,FALSE,TRUE)], collapse = " ")
questions[7] <- paste(possible[c(FALSE,FALSE,TRUE,TRUE)], collapse = " ")
questions[8] <- paste(possible[c(TRUE,FALSE,TRUE,TRUE)], collapse = " ")

solutions <- questions==paste(possible[trigger], collapse = " ")

explanations <- character(8)
explanations[1] <- ""
explanations[2] <- ""
explanations[3] <- ""
explanations[4] <- ""
explanations[5] <- ""
explanations[6] <- ""
explanations[7] <- ""
explanations[8] <- ""

## permute order of solutions/questions
Order <- sample(c(sample(which(solutions!=TRUE),4),which(solutions==TRUE)))
questions <- questions[Order]
solutions <- solutions[Order]
explanations <- explanations[Order]  
```    

Question
========

Download the data file [timeseries.csv](timeseries.csv) by right clicking and 'Save link as', then load it into RStudio using `read_csv()` or the Import Dataset button. 

In order to make sufficiently good predictions, which one of the following aspects has to be taken into your consideration when fitting a regression model to this time series?

```{r questionlist, echo = FALSE, results = "asis"}
answerlist(questions, markup = "markdown")
```

Solution
========

This question requires you to do a bit of exploratory analysis (i.e. plot the data):

```{r, message=FALSE, fig.caption = ""}
library(tidyverse)
d <- read_csv('timeseries.csv')
ggplot(data=d, mapping=aes(x=t,y=y)) +
  geom_line() +
  geom_point()
```

The plot may give you some ideas, but you can also explore fitting some models and doing diagnostics:

```{r, eval=FALSE}
# try some models with trend (linear, quadratic) and potential seasonality
m1 <- lm(y ~ t + s, data=d)
m2 <- lm(y ~ t + I(t^2) + s, data=d)
# try a transformation maybe?
m3 <- lm(log(y) ~ t + s, data=d)
m4 <- lm(log(y) ~ t + I(t^2) + s, data=d)

# see if terms are important:
anova(m1)

# try model diagnostics etc.
plot(m1)
```

```{r solutionlist, echo = FALSE, results = "asis"}
answerlist(mchoice2text(solutions, markup="markdown"), explanations, markup = "markdown")
```

Meta-information
================
extype: schoice
exsolution: `r mchoice2string(solutions)`
exname: Multiple Choice